---
title: "Generate COMPASS figures (platinum and NEPC endpoints)"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Canonical R figure workflow using tidyverse/ggplot2. **Knit** renders the
**ADT** cohort for both the **platinum** and **NEPC** endpoints by default.
ARPI remains available by explicit request.

The document reads the merged `profile_data` run under `.../CAIA/COMPASS`
and writes to `figures/CAIA/COMPASS`.

Outputs are split by cohort arm, then by endpoint, then flattened within each
figure directory. The platinum endpoint keeps the historical un-suffixed path
so existing references stay valid; NEPC nests one level deeper:

`FIG_ROOT/ADT/<figure>/adt_<plot-stem>.png`         (platinum)
`FIG_ROOT/ADT/nepc/<figure>/adt_<plot-stem>.png`    (NEPC)

For example, `figure1/adt_figure1a_consort.png` is written directly in
`figure1/`. Plot PDFs and legacy `original_*` Figure 2 variants are not
generated. Tables remain CSV/Markdown and use the same flat
figure/cohort-prefix convention. There is no `R/` language subdirectory.

Figure 2 (LLM classifier validation and platinum enrichment) describes the
label set and the platinum MRN list rather than a modelled endpoint, so it is
emitted once from the platinum pass and skipped for NEPC.

All figure-generation logic lives in `COMPASS_generate_figures_pipeline.R` as
`generate_figures(cohort, endpoint, ...)`. This document just sources that file
and calls it for each cohort in `COHORTS` (defaults to `c("adt")`) crossed with
each endpoint in `ENDPOINTS` (defaults to both).

Figure 4 also includes a supplemental held-out comparison of elastic-net Cox,
XGBoost, and Dynamic-DeepHit across mean AUC(t), C-index, and integrated Brier
score. Missing `03b_multivariate_longitudinal.ipynb` outputs are reported in
the figure and its data CSV rather than suppressing the supplement.

Run `01_preprocessing.ipynb` to completion first; this document only reads
that merged-profile run's outputs.

## Setup

```{r setup}
pipeline_path <- normalizePath(
  file.path(getwd(), "COMPASS_generate_figures_pipeline.R"),
  mustWork = FALSE
)
if (!file.exists(pipeline_path)) {
  pipeline_path <- normalizePath(
    file.path(getwd(), "COMPASS", "survival_analysis", "COMPASS_generate_figures_pipeline.R"),
    mustWork = FALSE
  )
}
if (!file.exists(pipeline_path))
  stop("Could not locate COMPASS_generate_figures_pipeline.R")
source(pipeline_path)

NEPC_PROJ_PATH <- "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS"
FIG_ROOT <- "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS"

# Endpoints to render. Each needs its own 01/02/03 build; see
# compass_pipeline.make_endpoint_runs().
ENDPOINTS <- c("platinum", "nepc")

# Generate lab-specific panels only for PSA and Testosterone.
# Non-androgen labs remain available to models and aggregate feature plots.
PLOT_NON_ANDROGEN_DISTRIBUTIONS <- FALSE
PLOT_NON_ANDROGEN_LAB_FIGURES <- FALSE

# Fail early with a useful message rather than deep inside generate_figures().
for (required_path in c(
    file.path(NEPC_PROJ_PATH, "LLM_NEPC_labels"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "platinum_MRN_list.csv"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "icd_prostate_mrn_flags.csv"),
    file.path(NEPC_PROJ_PATH, "longitudinal_prediction_data_adt.csv")
)) {
    if (!file.exists(required_path))
      stop(required_path, " not found -- run 01_preprocessing.ipynb first.")
}

# Each endpoint reads its own prediction-inputs tree. Check them up front so a
# missing NEPC build fails here rather than partway through the platinum pass.
for (COHORT in COHORTS) {
  for (ENDPOINT in ENDPOINTS) {
    suffix <- if (identical(ENDPOINT, "platinum")) "" else paste0("_", ENDPOINT)
    inputs_dir <- file.path(NEPC_PROJ_PATH, "survival_analysis",
                            paste0("prediction_inputs_", COHORT, suffix))
    if (!dir.exists(inputs_dir))
      stop(inputs_dir, " not found -- run 01_preprocessing.ipynb with ENDPOINTS = ",
           "(\"platinum\", \"nepc\") first.")
  }
}

message("data root:   ", NEPC_PROJ_PATH)
message("figure root: ", FIG_ROOT)
message("endpoints:   ", paste(ENDPOINTS, collapse = ", "))
```

## Render ADT figures

Runs the R figure-generation pipeline for ADT by default, once per endpoint.
Platinum outputs go to `FIG_ROOT/ADT/`, NEPC to `FIG_ROOT/ADT/nepc/`.
Set `COHORTS <- c("arpi", "adt")` to opt into both arms, or narrow `ENDPOINTS`
to render a single endpoint.

```{r render-figures}
for (COHORT in COHORTS) {
  for (ENDPOINT in ENDPOINTS) {
    message(sprintf("\n========== R figures: %s / %s ==========",
                    COHORT_LABELS[[COHORT]], toupper(ENDPOINT)))
    generate_figures(
      COHORT,
      nepc_proj_path = NEPC_PROJ_PATH,
      fig_root = FIG_ROOT,
      endpoint = ENDPOINT,
      plot_non_androgen_distributions = PLOT_NON_ANDROGEN_DISTRIBUTIONS,
      plot_non_androgen_lab_figures = PLOT_NON_ANDROGEN_LAB_FIGURES
    )
    endpoint_fig_root <- if (identical(ENDPOINT, "platinum")) {
      file.path(FIG_ROOT, toupper(COHORT))
    } else {
      file.path(FIG_ROOT, toupper(COHORT), ENDPOINT)
    }
    supplement_path <- file.path(
      endpoint_fig_root, "figure4",
      paste0(COHORT, "_figure4s_multivariate_all_models.png")
    )
    if (!file.exists(supplement_path))
      stop("Expected multivariate supplement was not generated: ", supplement_path)
    message("multivariate supplement: ", supplement_path)
    message(sprintf("\nCompleted R figures for %s / %s.",
                    COHORT_LABELS[[COHORT]], toupper(ENDPOINT)))
  }
}
```
