---
title: "Generate COMPASS platinum-arm figures"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Canonical R figure workflow using tidyverse/ggplot2. **Knit** renders the
**ADT** cohort by default. ARPI remains available by explicit request.

The document reads the merged `profile_data` run under `.../CAIA/COMPASS`
and writes to `figures/CAIA/COMPASS`.

Outputs are split by cohort arm, then flattened within each figure directory:

`FIG_ROOT/ADT/<figure>/adt_<plot-stem>.png`

For example, `figure1/adt_figure1a_consort.png` is written directly in
`figure1/`. Plot PDFs and legacy `original_*` Figure 2 variants are not
generated. Tables remain CSV/Markdown and use the same flat
figure/cohort-prefix convention. There is no `R/` language subdirectory.

All figure-generation logic lives in `COMPASS_generate_figures_pipeline.R` as
`generate_figures(cohort, ...)`. This document just sources that file and
calls it for each cohort in `COHORTS`, which defaults to `c("adt")`.

Run `01_preprocessing.ipynb` to completion first; this document only reads
that merged-profile run's outputs.

## Setup

```{r setup}
pipeline_path <- normalizePath(
  file.path(getwd(), "COMPASS_generate_figures_pipeline.R"),
  mustWork = FALSE
)
if (!file.exists(pipeline_path)) {
  pipeline_path <- normalizePath(
    file.path(getwd(), "COMPASS", "survival_analysis", "COMPASS_generate_figures_pipeline.R"),
    mustWork = FALSE
  )
}
if (!file.exists(pipeline_path))
  stop("Could not locate COMPASS_generate_figures_pipeline.R")
source(pipeline_path)

NEPC_PROJ_PATH <- "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS"
FIG_ROOT <- "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS"

# Generate lab-specific panels only for PSA and Testosterone.
# Non-androgen labs remain available to models and aggregate feature plots.
PLOT_NON_ANDROGEN_DISTRIBUTIONS <- FALSE
PLOT_NON_ANDROGEN_LAB_FIGURES <- FALSE

# Fail early with a useful message rather than deep inside generate_figures().
for (required_path in c(
    file.path(NEPC_PROJ_PATH, "LLM_NEPC_labels"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "platinum_MRN_list.csv"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "icd_prostate_mrn_flags.csv"),
    file.path(NEPC_PROJ_PATH, "longitudinal_prediction_data_adt.csv")
)) {
    if (!file.exists(required_path))
      stop(required_path, " not found -- run 01_preprocessing.ipynb first.")
}

message("data root:   ", NEPC_PROJ_PATH)
message("figure root: ", FIG_ROOT)
```

## Render ADT figures

Runs the R figure-generation pipeline for ADT by default. Outputs are written under
`FIG_ROOT/ADT/`. Set `COHORTS <- c("arpi", "adt")` to opt into both.

```{r render-figures}
for (COHORT in COHORTS) {
  message(sprintf("\n========== R figures: %s ==========", COHORT_LABELS[[COHORT]]))
  generate_figures(
    COHORT,
    nepc_proj_path = NEPC_PROJ_PATH,
    fig_root = FIG_ROOT,
    plot_non_androgen_distributions = PLOT_NON_ANDROGEN_DISTRIBUTIONS,
    plot_non_androgen_lab_figures = PLOT_NON_ANDROGEN_LAB_FIGURES
  )
  message(sprintf("\nCompleted R figures for %s.", COHORT_LABELS[[COHORT]]))
}
```
