---
title: "Generate COMPASS figures (NEPC and platinum endpoints)"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Canonical R figure workflow using tidyverse/ggplot2. **Knit** renders every
modelled **ADT** cohort -- the full arm, the ADT-intent metastatic stratum, and
the two LLM metastatic strata, each with and without the pre-ADT castrate
exclusion -- for the **NEPC** and **platinum** endpoints. ARPI remains
available by explicit request.

The AVPC endpoint and the localized-adjuvant cohort are retired; no such trees
are built. AVPC and multi-class `primary_label` fields may be read while
normalizing classifier inputs, but no figures are generated from them. Figure
stratifications are limited to platinum and binary NEPC status.

The document reads the merged `profile_data` run under `.../CAIA/COMPASS`
and writes to `figures/CAIA/COMPASS`.

The figure tree is **figure-major**, so comparing one panel across cohorts is a
single directory listing rather than twelve paths differing mid-string. Each
treatment arm keeps its own subtree, but endpoint, patient subset and exclusion
all live in the leaf filename:

`FIG_ROOT/<ARM>/by_figure/<group>/<trimmed-artifact-name>/<endpoint>__<subset>__<exclusion>.png`

Artifact names omit information already supplied by their parent folders. For
example, a `figure3_` prefix is removed inside `figure3/`, and a PSA
longitudinal artifact omits both `longitudinal_` and `_psa` inside
`labs/Androgen axis/PSA/longitudinal/`.

so one leaf holds all twelve cohort x endpoint panels, endpoint first so they
sort into two contiguous six-cohort blocks:

```
ADT/by_figure/figure3/univariate_landmark180/
    nepc__all__incl.png                 nepc__all__noprecastrate.png
    nepc__metastatic_adt__incl.png      nepc__metastatic_adt__noprecastrate.png
    nepc__metastatic_llm__incl.png      nepc__metastatic_llm__noprecastrate.png
    platinum__all__incl.png             platinum__all__noprecastrate.png
    platinum__metastatic_adt__incl.png  platinum__metastatic_adt__noprecastrate.png
    platinum__metastatic_llm__incl.png  platinum__metastatic_llm__noprecastrate.png
```

`by_figure/` is the sole output tree. The generator removes the retired
`by_cohort/` mirror left by older runs so downloads cannot dereference its
symlinks into duplicate files.

Plot PDFs are not generated. Tables remain CSV/Markdown and follow the same
leaf-naming convention. There is no `R/` language subdirectory.

Figure 2 v3 validates the binary `has_nepc` classifier call over every
ADT-exposed patient. It is emitted once from the canonical `adt` / platinum
pass and skipped for the other cells. AVPC confusion, multi-class subtype, and
AVPC/NEPC enrichment panels are not generated.

All outputs under `by_figure/labs/` are likewise emitted only by the platinum
pass. Their KM curves model time-to-platinum, their distributions split on
platinum status, and their trajectories are descriptive platinum/classifier
strata; a second NEPC-labelled copy would not represent a different endpoint
analysis. Each run removes legacy `nepc__*.png` lab leaves for its cohort.
Longitudinal summaries use 180-day bins from one year before through five
years after the treatment anchor, with day 0 as a hard boundary. A second
trajectory set is fit directly during plotting with `mgcv::bam`;
fREML and shrinkage select smoothness without depending on Python GAM outputs.
Both figure sets are emitted on the original scale and in log space. The log
versions transform each lab measurement with `log1p` before patient/bin
aggregation or GAM fitting and use an `_log` filename suffix.
Trajectory and GAM titles contain only the cohort definition and contributing
patient count; the lab, scale, and stratum remain identified by the axes,
legend, and artifact path.
For the ADT arm, three additional PSA/testosterone diagnostics quantify the
pre-ADT data supporting those trajectories: any and recent measurement
coverage, patient-level five-year measurement-count categories (including
zero), and zero-anchored 180-day-bin availability. Platinum panels use the
entire analysis-cohort stratum as their denominator; NEPC panels use all
classifier-labeled cohort patients, including those with no androgen lab. The
coverage diagnostics retain their five-year pre-ADT lookback independently of
the shorter trajectory window.

Figure 3b adds the somatic, Gleason, and PRS univariate associations as their
own plots, one per index analysis. Each is built on its own cohort and index
date (sequencing = specimen collection, Gleason = the score date nearest ADT
start, PRS = ADT start), so they are not merged into the Figure 3 lab
volcanoes. Gleason is drawn as a forest plot because it contributes a single
continuous feature; sequencing and PRS get a volcano plus a top-hits forest.
These analyses exist for the ADT arm at landmark 0 only, and a missing
`cox_somatic_gleason/` tree is reported and skipped.

All figure-generation logic lives in `COMPASS_generate_figures_pipeline.R` as
`generate_figures(cohort, endpoint, ...)`. This document just sources that file
and calls it for each cohort in `COHORTS` crossed with each endpoint in
`ENDPOINTS`.

Figure 4 also includes a supplemental held-out comparison of elastic-net Cox,
XGBoost, and Dynamic-DeepHit across mean AUC(t), C-index, and integrated Brier
score. Missing `03b_multivariate_longitudinal.ipynb` outputs are reported in
the figure and its data CSV rather than suppressing the supplement.

Run `01_preprocessing.ipynb` to completion first; this document only reads
that merged-profile run's outputs.

## Setup

```{r setup}
pipeline_path <- normalizePath(
  file.path(getwd(), "COMPASS_generate_figures_pipeline.R"),
  mustWork = FALSE
)
if (!file.exists(pipeline_path)) {
  pipeline_path <- normalizePath(
    file.path(getwd(), "COMPASS", "survival_analysis", "COMPASS_generate_figures_pipeline.R"),
    mustWork = FALSE
  )
}
if (!file.exists(pipeline_path))
  stop("Could not locate COMPASS_generate_figures_pipeline.R")
source(pipeline_path)

NEPC_PROJ_PATH <- "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS"
FIG_ROOT <- "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS"

# Endpoints to render. Each needs its own 01/02/03 build; see
# compass_pipeline.make_endpoint_runs(). Must be a subset of the pipeline's
# SUPPORTED_ENDPOINTS -- naming a retired endpoint here fails in the
# ENDPOINT_SUFFIXES lookup below rather than at generate_figures().
ENDPOINTS <- c("nepc", "platinum")

# Cohorts to render: every modelled ADT run label, matching the notebooks'
# COHORTS x EXCLUSIONS cross in compass_pipeline.DEFAULT_COHORTS /
# DEFAULT_EXCLUSIONS. Narrow this to iterate on a single tree.
COHORTS <- c(
  "adt",                 "adt_noprecastrate",
  "adt_metastatic_adt",  "adt_metastatic_adt_noprecastrate",
  "adt_metastatic_llm",  "adt_metastatic_llm_noprecastrate"
)

stopifnot(
  all(ENDPOINTS %in% SUPPORTED_ENDPOINTS),
  all(COHORTS %in% SUPPORTED_COHORTS)
)

# Generate lab-specific panels only for PSA and Testosterone.
# Non-androgen labs remain available to models and aggregate feature plots.
PLOT_NON_ANDROGEN_DISTRIBUTIONS <- FALSE
PLOT_NON_ANDROGEN_LAB_FIGURES <- FALSE

# Rendering profile. The default is optimized for iterative work: 300 dpi,
# composites instead of duplicate component+composite PNGs, and existing
# artifacts are retained. Longitudinal lab panels are always refreshed because
# their cohort membership and binning rules can change without changing their
# paths. Set COMPASS_RENDER_PROFILE=publication for the original 600-dpi,
# all-panels, forced-regeneration behavior.
RENDER_PROFILE <- tolower(Sys.getenv("COMPASS_RENDER_PROFILE", "fast"))
if (!RENDER_PROFILE %in% c("fast", "publication"))
  stop("COMPASS_RENDER_PROFILE must be 'fast' or 'publication'")
RENDER_DPI <- if (RENDER_PROFILE == "publication") 600 else 300
RENDER_OUTPUT_MODE <- if (RENDER_PROFILE == "publication") "all" else "composite"
overwrite_value <- tolower(Sys.getenv(
  "COMPASS_RENDER_OVERWRITE",
  if (RENDER_PROFILE == "publication") "true" else "false"
))
if (!overwrite_value %in% c("true", "false", "1", "0", "yes", "no"))
  stop("COMPASS_RENDER_OVERWRITE must be true or false")
RENDER_OVERWRITE <- overwrite_value %in% c("true", "1", "yes")

# Parallelize the cohort x endpoint cells directly so the full render can use
# more than the two endpoint slots. Override this default through
# COMPASS_RENDER_WORKERS when a lower-memory run is needed.
RENDER_WORKERS <- suppressWarnings(as.integer(
  Sys.getenv("COMPASS_RENDER_WORKERS", "12")
))
if (is.na(RENDER_WORKERS) || RENDER_WORKERS < 1L)
  stop("COMPASS_RENDER_WORKERS must be a positive integer")

# Fail early with a useful message rather than deep inside generate_figures().
for (required_path in c(
    file.path(NEPC_PROJ_PATH, "LLM_NEPC_labels"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "platinum_MRN_list.csv"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "icd_prostate_mrn_flags.csv"),
    file.path(NEPC_PROJ_PATH, "longitudinal_prediction_data_adt.csv")
)) {
    if (!file.exists(required_path))
      stop(required_path, " not found -- run 01_preprocessing.ipynb first.")
}

# Each cohort x endpoint reads its own prediction-inputs tree. Check them all up
# front and report EVERY missing one together: with 6 cohorts x 2 endpoints,
# stopping at the first would mean 12 edit-rerun cycles to discover an
# incomplete 01_preprocessing pass.
missing_trees <- character(0)
for (COHORT in COHORTS) {
  for (ENDPOINT in ENDPOINTS) {
    suffix <- unname(ENDPOINT_SUFFIXES[[ENDPOINT]])
    inputs_dir <- file.path(NEPC_PROJ_PATH, "survival_analysis",
                            paste0("prediction_inputs_", COHORT, suffix))
    if (!dir.exists(inputs_dir))
      missing_trees <- c(missing_trees, inputs_dir)
  }
}
if (length(missing_trees))
  stop(length(missing_trees), " of ", length(COHORTS) * length(ENDPOINTS),
       " prediction-input trees are missing -- run 01_preprocessing.ipynb for ",
       "every cohort x endpoint first:\n  ",
       paste(missing_trees, collapse = "\n  "))

message("data root:   ", NEPC_PROJ_PATH)
message("figure root: ", FIG_ROOT)
message("endpoints:   ", paste(ENDPOINTS, collapse = ", "))
message("cohorts:     ", paste(COHORTS, collapse = ", "))
message(sprintf("renders:     %d cohort x endpoint figure sets",
                length(COHORTS) * length(ENDPOINTS)))
message(sprintf("profile:     %s (%d dpi, output=%s, overwrite=%s, workers=%d)",
                RENDER_PROFILE, RENDER_DPI, RENDER_OUTPUT_MODE,
                RENDER_OVERWRITE, RENDER_WORKERS))
```

## Render ADT figures

Runs the R figure-generation pipeline once per cohort x endpoint -- 6 cohorts x
2 endpoints = 12 figure sets. Outputs go to `FIG_ROOT/<ARM>/by_figure/`, with
each run writing one `<endpoint>__<subset>__<exclusion>` leaf per plot stem.

Narrow `COHORTS` or `ENDPOINTS` above to iterate on a single tree. `COHORTS`
accepts any run label in `SUPPORTED_COHORTS`, so the ARPI arm is rendered by
naming its labels, and a retired tree that still exists on disk (an
`adt_localized` or `_avpc` build) can be re-rendered by naming it explicitly.

Each cohort x endpoint cell is rendered independently, with up to twelve cells
running concurrently. A failure in one is reported rather than abandoning the
other 11. Any failures are re-reported at the end and the chunk then fails, so
a partial render cannot pass silently.

```{r render-figures}
render_cell <- function(COHORT, ENDPOINT) {
  label <- sprintf("%s / %s", cohort_display(COHORT), toupper(ENDPOINT))
  message(sprintf("\n========== R figures: %s ==========", label))
  ok <- tryCatch({
    generate_figures(
      COHORT,
      nepc_proj_path = NEPC_PROJ_PATH,
      fig_root = FIG_ROOT,
      endpoint = ENDPOINT,
      plot_non_androgen_distributions = PLOT_NON_ANDROGEN_DISTRIBUTIONS,
      plot_non_androgen_lab_figures = PLOT_NON_ANDROGEN_LAB_FIGURES,
      save_dpi = RENDER_DPI,
      output_mode = RENDER_OUTPUT_MODE,
      overwrite = RENDER_OVERWRITE
    )
    supplement_path <- file.path(
      FIG_ROOT, toupper(cohort_arm(COHORT)), "by_figure", "figure4",
      "s_multivariate_all_models",
      sprintf("%s__%s.png", ENDPOINT, cohort_leaf_slug(COHORT))
    )
    # The supplement is the last artifact generate_figures() writes, so its
    # absence means the pass did not complete even when nothing threw.
    if (!file.exists(supplement_path))
      stop("expected multivariate supplement was not generated: ",
           supplement_path)
    message("multivariate supplement: ", supplement_path)
    TRUE
  }, error = function(e) {
    message(sprintf("FAILED %s: %s", label, conditionMessage(e)))
    FALSE
  })
  if (ok) {
    message(sprintf("\nCompleted R figures for %s.", label))
    character(0)
  } else {
    label
  }
}

render_grid <- tidyr::crossing(COHORT = COHORTS, ENDPOINT = ENDPOINTS)
render_one <- function(i) {
  render_cell(render_grid$COHORT[[i]], render_grid$ENDPOINT[[i]])
}
n_workers <- min(RENDER_WORKERS, nrow(render_grid))
if (n_workers > 1L && .Platform$OS.type != "windows") {
  render_results <- parallel::mclapply(
    seq_len(nrow(render_grid)), render_one,
    mc.cores = n_workers, mc.preschedule = TRUE
  )
} else {
  if (n_workers > 1L)
    message("Parallel fork rendering is unavailable on Windows; using one worker.")
  render_results <- lapply(seq_len(nrow(render_grid)), render_one)
}
worker_errors <- vapply(render_results, inherits, logical(1), what = "try-error")
render_failures <- unlist(render_results[!worker_errors], use.names = FALSE)
if (any(worker_errors))
  render_failures <- c(
    render_failures,
    unlist(lapply(which(worker_errors), function(i) {
      sprintf("%s / %s (worker failed)", cohort_display(render_grid$COHORT[[i]]),
              toupper(render_grid$ENDPOINT[[i]]))
    }), use.names = FALSE)
  )

n_total <- length(COHORTS) * length(ENDPOINTS)
message(sprintf("\n%d of %d figure sets rendered.",
                n_total - length(render_failures), n_total))
if (length(render_failures))
  stop(length(render_failures), " of ", n_total, " figure sets failed:\n  ",
       paste(render_failures, collapse = "\n  "))
```
