---
title: "Generate COMPASS figures (NEPC and platinum endpoints)"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Canonical R figure workflow using tidyverse/ggplot2. **Knit** renders every
modelled **ADT** cohort -- the full arm, the ADT-intent metastatic stratum, and
the two LLM metastatic strata, each with and without the pre-ADT castrate
exclusion -- for the **NEPC** and **platinum** endpoints. ARPI remains
available by explicit request.

The AVPC endpoint and the localized-adjuvant cohort are retired; no such trees
are built. AVPC and multi-class `primary_label` fields may be read while
normalizing classifier inputs, but no figures are generated from them. Figure
stratifications are limited to platinum and binary NEPC status.

The document reads the merged `profile_data` run under `.../CAIA/COMPASS`
and writes to `figures/CAIA/COMPASS`.

The figure tree is **figure-major**, so comparing one panel across cohorts is a
single directory listing rather than twelve paths differing mid-string. Each
treatment arm keeps its own subtree, but endpoint, patient subset and exclusion
all live in the leaf filename:

`FIG_ROOT/<ARM>/by_figure/<group>/<trimmed-artifact-name>/<endpoint>__<subset>__<exclusion>.png`

Artifact names omit information already supplied by their parent folders. For
example, a `figure3_` prefix is removed inside `figure3/`, and a PSA
longitudinal artifact omits both `longitudinal_` and `_psa` inside
`labs/Androgen axis/PSA/longitudinal/`.

so one leaf holds all twelve cohort x endpoint panels, endpoint first so they
sort into two contiguous six-cohort blocks:

```
ADT/by_figure/figure3/univariate_landmark180/
    nepc__all__incl.png                 nepc__all__noprecastrate.png
    nepc__metastatic_adt__incl.png      nepc__metastatic_adt__noprecastrate.png
    nepc__metastatic_llm__incl.png      nepc__metastatic_llm__noprecastrate.png
    platinum__all__incl.png             platinum__all__noprecastrate.png
    platinum__metastatic_adt__incl.png  platinum__metastatic_adt__noprecastrate.png
    platinum__metastatic_llm__incl.png  platinum__metastatic_llm__noprecastrate.png
```

`by_figure/` is the sole output tree. The generator removes the retired
`by_cohort/` mirror left by older runs so downloads cannot dereference its
symlinks into duplicate files.

PNG is the default and always-rendered figure format. Set
`COMPASS_RENDER_PDF=true` to additionally write a vector PDF beside every PNG;
when false, stale PDFs are removed. Tables remain CSV/Markdown and follow the
same leaf-naming convention. There is no `R/` language subdirectory.

Figure 2 v3 validates the binary `has_nepc` classifier call over every
ADT-exposed patient. It is emitted once from the canonical `adt` / platinum
pass and skipped for the other cells. AVPC confusion, multi-class subtype, and
AVPC/NEPC enrichment panels are not generated.

All outputs under `by_figure/labs/` are likewise emitted only by the platinum
pass. Lab outputs are limited to longitudinal summaries; R-fitted GAM panels
remain available but are toggled off by default. The distribution and
quartile-KM families are retired. A second NEPC-labelled copy would not
represent a different endpoint analysis. Each run removes legacy
`nepc__*.png` lab leaves for its cohort.
Longitudinal summaries use 180-day bins from one year before through five
years after the treatment anchor, with day 0 as a hard boundary. A second
trajectory set can be fit directly during plotting with `mgcv::bam` when
`PLOT_GAM_TRAJECTORIES` is enabled; fREML and shrinkage select smoothness
without depending on Python GAM outputs. Both implementations support the
original scale and log space. The log versions transform each lab measurement
with `log1p` before patient/bin aggregation or GAM fitting and use an `_log`
filename suffix.
Trajectory and GAM titles contain only the cohort definition and contributing
patient count; the lab, scale, and stratum remain identified by the axes,
legend, and artifact path.
For the ADT arm, two additional PSA/testosterone diagnostics quantify the
pre-ADT data supporting those trajectories: any/recent measurement coverage
and zero-anchored 180-day-bin availability. Platinum panels use the
entire analysis-cohort stratum as their denominator; NEPC panels use all
classifier-labeled cohort patients, including those with no androgen lab. The
coverage diagnostics retain their five-year pre-ADT lookback independently of
the shorter trajectory window.

Figure 3b is limited to sequencing-based somatic associations, shown as a
volcano plus a top-hits forest. PRS and Gleason panels are retired. The
sequencing analysis exists for the ADT arm at landmark 0 only, and a missing
`cox_somatic_gleason/` tree is reported and skipped.

All figure-generation logic lives in `COMPASS_generate_figures_pipeline.R` as
`generate_figures(cohort, endpoint, ...)`. This document just sources that file
and calls it for each cohort in `COHORTS` crossed with each endpoint in
`ENDPOINTS`.

The Figure 1 analysis-set supplement, Figure 3 significance-only variants,
Figure 4 all-model supplement, and LLM-stratified KM family are retired.

Run `01_preprocessing.ipynb` to completion first; this document only reads
that merged-profile run's outputs.

## Setup

```{r setup}
pipeline_path <- normalizePath(
  file.path(getwd(), "COMPASS_generate_figures_pipeline.R"),
  mustWork = FALSE
)
if (!file.exists(pipeline_path)) {
  pipeline_path <- normalizePath(
    file.path(getwd(), "COMPASS", "survival_analysis", "COMPASS_generate_figures_pipeline.R"),
    mustWork = FALSE
  )
}
if (!file.exists(pipeline_path))
  stop("Could not locate COMPASS_generate_figures_pipeline.R")
source(pipeline_path)

NEPC_PROJ_PATH <- "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS"
FIG_ROOT <- "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS"

# Endpoints to render. Each needs its own 01/02/03 build; see
# compass_pipeline.make_endpoint_runs(). Must be a subset of the pipeline's
# SUPPORTED_ENDPOINTS -- naming a retired endpoint here fails in the
# ENDPOINT_SUFFIXES lookup below rather than at generate_figures().
ENDPOINTS <- c("nepc", "platinum")

# Cohorts to render: every modelled ADT run label, matching the notebooks'
# COHORTS x EXCLUSIONS cross in compass_pipeline.DEFAULT_COHORTS /
# DEFAULT_EXCLUSIONS. Narrow this to iterate on a single tree.
COHORTS <- c(
  "adt",                 "adt_noprecastrate",
  "adt_metastatic_adt",  "adt_metastatic_adt_noprecastrate",
  "adt_metastatic_llm",  "adt_metastatic_llm_noprecastrate"
)

stopifnot(
  all(ENDPOINTS %in% SUPPORTED_ENDPOINTS),
  all(COHORTS %in% SUPPORTED_COHORTS)
)

# Generate lab-specific panels only for PSA and Testosterone.
# Non-androgen labs remain available to models and aggregate feature plots.
PLOT_NON_ANDROGEN_DISTRIBUTIONS <- FALSE
PLOT_NON_ANDROGEN_LAB_FIGURES <- FALSE
# Keep the R/mgcv fitting implementation available without running its
# comparatively expensive trajectory fits during the default figure knit.
PLOT_GAM_TRAJECTORIES <- FALSE

# Rendering profile. The default is optimized for iterative work: 300 dpi,
# composites instead of duplicate component+composite PNGs, and existing
# artifacts are retained. Longitudinal lab panels are always refreshed because
# their cohort membership and binning rules can change without changing their
# paths. Set COMPASS_RENDER_PROFILE=publication for the original 600-dpi,
# all-panels, forced-regeneration behavior.
RENDER_PROFILE <- tolower(Sys.getenv("COMPASS_RENDER_PROFILE", "fast"))
if (!RENDER_PROFILE %in% c("fast", "publication"))
  stop("COMPASS_RENDER_PROFILE must be 'fast' or 'publication'")
RENDER_DPI <- if (RENDER_PROFILE == "publication") 600 else 300
RENDER_OUTPUT_MODE <- if (RENDER_PROFILE == "publication") "all" else "composite"
overwrite_value <- tolower(Sys.getenv(
  "COMPASS_RENDER_OVERWRITE",
  if (RENDER_PROFILE == "publication") "true" else "false"
))
if (!overwrite_value %in% c("true", "false", "1", "0", "yes", "no"))
  stop("COMPASS_RENDER_OVERWRITE must be true or false")
RENDER_OVERWRITE <- overwrite_value %in% c("true", "1", "yes")

pdf_value <- tolower(Sys.getenv("COMPASS_RENDER_PDF", "false"))
if (!pdf_value %in% c("true", "false", "1", "0", "yes", "no"))
  stop("COMPASS_RENDER_PDF must be true or false")
RENDER_PDF <- pdf_value %in% c("true", "1", "yes")

# Parallelize cohort x endpoint cells, but keep the default conservative: each
# render holds large plotting data and GAM fits, so launching all twelve cells
# together can exhaust memory and make the knit appear to freeze. Increase the
# worker count explicitly only on a host with enough RAM.
RENDER_WORKERS <- suppressWarnings(as.integer(
  Sys.getenv("COMPASS_RENDER_WORKERS", "2")
))
if (is.na(RENDER_WORKERS) || RENDER_WORKERS < 1L)
  stop("COMPASS_RENDER_WORKERS must be a positive integer")

# Fail early with a useful message rather than deep inside generate_figures().
for (required_path in c(
    file.path(NEPC_PROJ_PATH, "LLM_NEPC_labels"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "platinum_MRN_list.csv"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "icd_prostate_mrn_flags.csv"),
    file.path(NEPC_PROJ_PATH, "longitudinal_prediction_data_adt.csv")
)) {
    if (!file.exists(required_path))
      stop(required_path, " not found -- run 01_preprocessing.ipynb first.")
}

# Each cohort x endpoint reads its own prediction-inputs tree. Check them all up
# front and report EVERY missing one together: with 6 cohorts x 2 endpoints,
# stopping at the first would mean 12 edit-rerun cycles to discover an
# incomplete 01_preprocessing pass.
missing_trees <- character(0)
for (COHORT in COHORTS) {
  for (ENDPOINT in ENDPOINTS) {
    suffix <- unname(ENDPOINT_SUFFIXES[[ENDPOINT]])
    inputs_dir <- file.path(NEPC_PROJ_PATH, "survival_analysis",
                            paste0("prediction_inputs_", COHORT, suffix))
    if (!dir.exists(inputs_dir))
      missing_trees <- c(missing_trees, inputs_dir)
  }
}
if (length(missing_trees))
  stop(length(missing_trees), " of ", length(COHORTS) * length(ENDPOINTS),
       " prediction-input trees are missing -- run 01_preprocessing.ipynb for ",
       "every cohort x endpoint first:\n  ",
       paste(missing_trees, collapse = "\n  "))

# Populate arm-level caches before parallel::mclapply forks. On Unix, workers
# then inherit these large read-only objects through copy-on-write memory
# instead of parsing the longitudinal CSV independently up to 12 times.
# Sequential/Windows renders benefit from the same warm caches.
shared_longitudinal_csv <- file.path(
  NEPC_PROJ_PATH, "longitudinal_prediction_data_adt.csv"
)
preload_started <- proc.time()[["elapsed"]]
invisible(cached_profile_patient_and_labs(shared_longitudinal_csv, "DFCI_MRN"))
invisible(cached_canonical_longitudinal(shared_longitudinal_csv))

# These sources are smaller, but every cell reads them. Warm their raw caches
# as well so forked workers do no repeated shared-input I/O.
invisible(cached_read_csv(
  file.path(NEPC_PROJ_PATH, "LLM_NEPC_labels", "baca_lab_annotations.csv"),
  show_col_types = FALSE
))
invisible(cached_read_csv(
  file.path(NEPC_PROJ_PATH, "mrn_lists", "platinum_MRN_list.csv"),
  show_col_types = FALSE
))
invisible(cached_read_csv(
  file.path(NEPC_PROJ_PATH, "mrn_lists", "icd_prostate_mrn_flags.csv"),
  show_col_types = FALSE
))
invisible(load_llm_strata(DEFAULT_LLM_ANNOTATIONS_PATH))
message(sprintf(
  "shared inputs preloaded once in %.1f seconds; forking workers from warm cache",
  proc.time()[["elapsed"]] - preload_started
))

message("data root:   ", NEPC_PROJ_PATH)
message("figure root: ", FIG_ROOT)
message("endpoints:   ", paste(ENDPOINTS, collapse = ", "))
message("cohorts:     ", paste(COHORTS, collapse = ", "))
message(sprintf("renders:     %d cohort x endpoint figure sets",
                length(COHORTS) * length(ENDPOINTS)))
message(sprintf("profile:     %s (%d dpi, output=%s, overwrite=%s, pdf=%s, workers=%d)",
                RENDER_PROFILE, RENDER_DPI, RENDER_OUTPUT_MODE,
                RENDER_OVERWRITE, RENDER_PDF, RENDER_WORKERS))
```

## Render ADT figures

Runs the R figure-generation pipeline once per cohort x endpoint -- 6 cohorts x
2 endpoints = 12 figure sets. Outputs go to `FIG_ROOT/<ARM>/by_figure/`, with
each run writing one `<endpoint>__<subset>__<exclusion>` leaf per plot stem.

Narrow `COHORTS` or `ENDPOINTS` above to iterate on a single tree. `COHORTS`
accepts any run label in `SUPPORTED_COHORTS`, so the ARPI arm is rendered by
naming its labels, and a retired tree that still exists on disk (an
`adt_localized` or `_avpc` build) can be re-rendered by naming it explicitly.

Each cohort x endpoint cell is rendered independently, with two cells running
concurrently by default. A failure in one is reported rather than abandoning
the other 11. Any failures are re-reported at the end and the chunk then fails,
so a partial render cannot pass silently.

```{r render-figures}
render_cell <- function(COHORT, ENDPOINT) {
  label <- sprintf("%s / %s", cohort_display(COHORT), toupper(ENDPOINT))
  cell_started <- proc.time()[["elapsed"]]
  message(sprintf("\n========== R figures: %s ==========", label))
  ok <- tryCatch({
    generate_figures(
      COHORT,
      nepc_proj_path = NEPC_PROJ_PATH,
      fig_root = FIG_ROOT,
      endpoint = ENDPOINT,
      plot_non_androgen_distributions = PLOT_NON_ANDROGEN_DISTRIBUTIONS,
      plot_non_androgen_lab_figures = PLOT_NON_ANDROGEN_LAB_FIGURES,
      plot_gam_trajectories = PLOT_GAM_TRAJECTORIES,
      save_dpi = RENDER_DPI,
      save_pdf = RENDER_PDF,
      output_mode = RENDER_OUTPUT_MODE,
      overwrite = RENDER_OVERWRITE
    )
    completion_path <- file.path(
      FIG_ROOT, toupper(cohort_arm(COHORT)), "by_figure", "figure4",
      "multivariate_performance",
      sprintf("%s__%s.png", ENDPOINT, cohort_leaf_slug(COHORT))
    )
    # Confirm that the core composite was emitted even when nothing threw.
    if (!file.exists(completion_path))
      stop("expected multivariate figure was not generated: ", completion_path)
    message("multivariate figure: ", completion_path)
    TRUE
  }, error = function(e) {
    message(sprintf("FAILED %s: %s", label, conditionMessage(e)))
    FALSE
  })
  if (ok) {
    message(sprintf("\nCompleted R figures for %s in %.1f seconds.", label,
                    proc.time()[["elapsed"]] - cell_started))
    character(0)
  } else {
    label
  }
}

render_grid <- tidyr::crossing(COHORT = COHORTS, ENDPOINT = ENDPOINTS)
render_one <- function(i) {
  render_cell(render_grid$COHORT[[i]], render_grid$ENDPOINT[[i]])
}
n_workers <- min(RENDER_WORKERS, nrow(render_grid))
if (n_workers > 1L && .Platform$OS.type != "windows") {
  render_results <- parallel::mclapply(
    seq_len(nrow(render_grid)), render_one,
    mc.cores = n_workers, mc.preschedule = TRUE
  )
} else {
  if (n_workers > 1L)
    message("Parallel fork rendering is unavailable on Windows; using one worker.")
  render_results <- lapply(seq_len(nrow(render_grid)), render_one)
}
worker_errors <- vapply(render_results, inherits, logical(1), what = "try-error")
render_failures <- unlist(render_results[!worker_errors], use.names = FALSE)
if (any(worker_errors))
  render_failures <- c(
    render_failures,
    unlist(lapply(which(worker_errors), function(i) {
      sprintf("%s / %s (worker failed)", cohort_display(render_grid$COHORT[[i]]),
              toupper(render_grid$ENDPOINT[[i]]))
    }), use.names = FALSE)
  )

n_total <- length(COHORTS) * length(ENDPOINTS)
message(sprintf("\n%d of %d figure sets rendered.",
                n_total - length(render_failures), n_total))
if (length(render_failures))
  stop(length(render_failures), " of ", n_total, " figure sets failed:\n  ",
       paste(render_failures, collapse = "\n  "))
```
