---
title: "Generate COMPASS figures (NEPC and platinum endpoints)"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Canonical R figure workflow using tidyverse/ggplot2. **Knit** renders every
modelled **ADT** cohort -- the full arm, the ADT-intent metastatic stratum, and
the two LLM metastatic strata, each with and without the pre-ADT castrate
exclusion -- for the **NEPC** and **platinum** endpoints. ARPI remains
available by explicit request.

The AVPC endpoint and the localized-adjuvant cohort are retired; no such trees
are built. AVPC survives elsewhere as a *phenotype label* (`has_avpc`,
`primary_label`) in Figure 2's classifier validation, which is a different
thing from the retired survival endpoint.

The document reads the merged `profile_data` run under `.../CAIA/COMPASS`
and writes to `figures/CAIA/COMPASS`.

Outputs are split by cohort arm, then by endpoint, then flattened within each
figure directory. Every endpoint nests one level below its cohort arm --
platinum included, so no endpoint's panels can overwrite another's:

`FIG_ROOT/ADT/nepc/<figure>/adt_<plot-stem>.png`
`FIG_ROOT/ADT/platinum/<figure>/adt_<plot-stem>.png`
`FIG_ROOT/ADT_LLM_METASTATIC/nepc/<figure>/adt_llm_metastatic_<plot-stem>.png`

The cohort directory is the run label upper-cased, so each of the eight run
labels gets its own tree and no two cohorts can overwrite each other.

Note that platinum figures moved out of the old un-suffixed `FIG_ROOT/ADT/`
path; references to that location need updating. For example,
`figure1/adt_figure1a_consort.png` is written directly in `platinum/figure1/`.
Plot PDFs are not generated. Tables remain CSV/Markdown and use the same flat
figure/cohort-prefix convention. There is no `R/` language subdirectory.

Figure 2 v3 (LLM classifier validation and platinum enrichment over every
ADT-exposed patient) describes the label set and the platinum MRN list rather
than a modelled endpoint, so it is emitted once from the platinum pass and
skipped for the other endpoints. It is the only Figure 2 variant retained;
the earlier v0/v1/v2 variants have been removed.

Figure 3b adds the somatic, Gleason, and PRS univariate associations as their
own plots, one per index analysis. Each is built on its own cohort and index
date (sequencing = specimen collection, Gleason = the score date nearest ADT
start, PRS = ADT start), so they are not merged into the Figure 3 lab
volcanoes. Gleason is drawn as a forest plot because it contributes a single
continuous feature; sequencing and PRS get a volcano plus a top-hits forest.
These analyses exist for the ADT arm at landmark 0 only, and a missing
`cox_somatic_gleason/` tree is reported and skipped.

All figure-generation logic lives in `COMPASS_generate_figures_pipeline.R` as
`generate_figures(cohort, endpoint, ...)`. This document just sources that file
and calls it for each cohort in `COHORTS` crossed with each endpoint in
`ENDPOINTS`.

Figure 4 also includes a supplemental held-out comparison of elastic-net Cox,
XGBoost, and Dynamic-DeepHit across mean AUC(t), C-index, and integrated Brier
score. Missing `03b_multivariate_longitudinal.ipynb` outputs are reported in
the figure and its data CSV rather than suppressing the supplement.

Run `01_preprocessing.ipynb` to completion first; this document only reads
that merged-profile run's outputs.

## Setup

```{r setup}
pipeline_path <- normalizePath(
  file.path(getwd(), "COMPASS_generate_figures_pipeline.R"),
  mustWork = FALSE
)
if (!file.exists(pipeline_path)) {
  pipeline_path <- normalizePath(
    file.path(getwd(), "COMPASS", "survival_analysis", "COMPASS_generate_figures_pipeline.R"),
    mustWork = FALSE
  )
}
if (!file.exists(pipeline_path))
  stop("Could not locate COMPASS_generate_figures_pipeline.R")
source(pipeline_path)

NEPC_PROJ_PATH <- "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS"
FIG_ROOT <- "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS"

# Endpoints to render. Each needs its own 01/02/03 build; see
# compass_pipeline.make_endpoint_runs(). Must be a subset of the pipeline's
# SUPPORTED_ENDPOINTS -- naming a retired endpoint here fails in the
# ENDPOINT_SUFFIXES lookup below rather than at generate_figures().
ENDPOINTS <- c("nepc", "platinum")

# Cohorts to render: every modelled ADT run label, matching the notebooks'
# COHORTS x EXCLUSIONS cross in compass_pipeline.DEFAULT_COHORTS /
# DEFAULT_EXCLUSIONS. Narrow this to iterate on a single tree.
COHORTS <- c(
  "adt",                    "adt_noprecastrate",
  "adt_metastatic",         "adt_metastatic_noprecastrate",
  "adt_llm_metastatic",     "adt_llm_metastatic_noprecastrate",
  "adt_llm_nonmetastatic",  "adt_llm_nonmetastatic_noprecastrate"
)

stopifnot(
  all(ENDPOINTS %in% SUPPORTED_ENDPOINTS),
  all(COHORTS %in% SUPPORTED_COHORTS)
)

# Generate lab-specific panels only for PSA and Testosterone.
# Non-androgen labs remain available to models and aggregate feature plots.
PLOT_NON_ANDROGEN_DISTRIBUTIONS <- FALSE
PLOT_NON_ANDROGEN_LAB_FIGURES <- FALSE

# Fail early with a useful message rather than deep inside generate_figures().
for (required_path in c(
    file.path(NEPC_PROJ_PATH, "LLM_NEPC_labels"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "platinum_MRN_list.csv"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "icd_prostate_mrn_flags.csv"),
    file.path(NEPC_PROJ_PATH, "longitudinal_prediction_data_adt.csv")
)) {
    if (!file.exists(required_path))
      stop(required_path, " not found -- run 01_preprocessing.ipynb first.")
}

# Each cohort x endpoint reads its own prediction-inputs tree. Check them all up
# front and report EVERY missing one together: with 8 cohorts x 2 endpoints,
# stopping at the first would mean 16 edit-rerun cycles to discover an
# incomplete 01_preprocessing pass.
missing_trees <- character(0)
for (COHORT in COHORTS) {
  for (ENDPOINT in ENDPOINTS) {
    suffix <- unname(ENDPOINT_SUFFIXES[[ENDPOINT]])
    inputs_dir <- file.path(NEPC_PROJ_PATH, "survival_analysis",
                            paste0("prediction_inputs_", COHORT, suffix))
    if (!dir.exists(inputs_dir))
      missing_trees <- c(missing_trees, inputs_dir)
  }
}
if (length(missing_trees))
  stop(length(missing_trees), " of ", length(COHORTS) * length(ENDPOINTS),
       " prediction-input trees are missing -- run 01_preprocessing.ipynb for ",
       "every cohort x endpoint first:\n  ",
       paste(missing_trees, collapse = "\n  "))

message("data root:   ", NEPC_PROJ_PATH)
message("figure root: ", FIG_ROOT)
message("endpoints:   ", paste(ENDPOINTS, collapse = ", "))
message("cohorts:     ", paste(COHORTS, collapse = ", "))
message(sprintf("renders:     %d cohort x endpoint figure sets",
                length(COHORTS) * length(ENDPOINTS)))
```

## Render ADT figures

Runs the R figure-generation pipeline once per cohort x endpoint -- 8 cohorts x
2 endpoints = 16 figure sets. Outputs go to `FIG_ROOT/<COHORT>/<endpoint>/`,
e.g. `ADT/platinum/`, `ADT_LLM_METASTATIC/nepc/`.

Narrow `COHORTS` or `ENDPOINTS` above to iterate on a single tree. `COHORTS`
accepts any run label in `SUPPORTED_COHORTS`, so the ARPI arm is rendered by
naming its labels, and a retired tree that still exists on disk (an
`adt_localized` or `_avpc` build) can be re-rendered by naming it explicitly.

Each cohort is rendered independently, and a failure in one is reported and
skipped rather than abandoning the other 15 -- a single cohort with too few
events should not cost the whole knit. Any failures are re-reported at the end
and the chunk then fails, so a partial render cannot pass silently.

```{r render-figures}
render_failures <- character(0)

for (COHORT in COHORTS) {
  for (ENDPOINT in ENDPOINTS) {
    label <- sprintf("%s / %s", cohort_display(COHORT), toupper(ENDPOINT))
    message(sprintf("\n========== R figures: %s ==========", label))
    ok <- tryCatch({
      generate_figures(
        COHORT,
        nepc_proj_path = NEPC_PROJ_PATH,
        fig_root = FIG_ROOT,
        endpoint = ENDPOINT,
        plot_non_androgen_distributions = PLOT_NON_ANDROGEN_DISTRIBUTIONS,
        plot_non_androgen_lab_figures = PLOT_NON_ANDROGEN_LAB_FIGURES
      )
      endpoint_fig_root <- file.path(FIG_ROOT, toupper(COHORT), ENDPOINT)
      supplement_path <- file.path(
        endpoint_fig_root, "figure4",
        paste0(COHORT, "_figure4s_multivariate_all_models.png")
      )
      # The supplement is the last artifact generate_figures() writes, so its
      # absence means the pass did not complete even when nothing threw.
      if (!file.exists(supplement_path))
        stop("expected multivariate supplement was not generated: ",
             supplement_path)
      message("multivariate supplement: ", supplement_path)
      TRUE
    }, error = function(e) {
      message(sprintf("FAILED %s: %s", label, conditionMessage(e)))
      FALSE
    })
    if (ok) {
      message(sprintf("\nCompleted R figures for %s.", label))
    } else {
      render_failures <- c(render_failures, label)
    }
  }
}

n_total <- length(COHORTS) * length(ENDPOINTS)
message(sprintf("\n%d of %d figure sets rendered.",
                n_total - length(render_failures), n_total))
if (length(render_failures))
  stop(length(render_failures), " of ", n_total, " figure sets failed:\n  ",
       paste(render_failures, collapse = "\n  "))
```
