---
title: "COMPASS figures: cached preparation and rendering"
output:
  html_document:
    toc: true
    toc_depth: 2
---

## Workflow

Run `04_prep_figure_data.ipynb` first to prepare the Polars tables and manifest.
All Python execution belongs to that notebook. This document and its R helpers
never invoke Python: they consume the notebook's prepared data only. Missing,
changed, or incompatible inputs stop with instructions to rerun the notebook.

Preparation and rendering are separate stages:

1. `04_prep_figure_data.ipynb` uses Polars to scan each selected arm's longitudinal
   CSV once and cache patient metadata, all-lab record spans, selected lab
   measurements, patient/bin raw and log means, and pre-treatment coverage.
2. R consumes those tables, calculates the existing survival statistics and
   optional GAMs, and caches finished graphics objects. CSV/Markdown figure
   tables are written during this stage; no images are generated.
3. Rendering loads the graphics objects and writes individual PNGs and optional
   PDFs. It does not read patient inputs, fit curves, or repeat plot calculations.

`COMPASS_FIGURE_STAGE=all` (default) checks the notebook manifest and source
metadata, prepares changed R figure sets, then renders missing or stale formats.
`prepare` builds R graphics caches only; it does not prepare Python tables.
`render` uses the last prepared snapshot without checking
source data, and requires the same cohort/analysis settings. Use `all` for
normal updates and `render` for DPI/PDF changes.

`COMPASS_FIGURE_SCOPE=federated` processes only the no-MSK comparison without
requiring longitudinal data or running other figure families.

Caches live under `COMPASS_FIGURE_DATA_ROOT` (default `COMPASS/figure_data`).
They include patient data and need the same access restrictions as the analysis
inputs. Versioned caches are retained. Source identity uses absolute paths,
file sizes and nanosecond modification times, plus code/settings versions.
Use `FORCE = True` in `04_prep_figure_data.ipynb` after replacing sources while
preserving timestamps and sizes. In this Rmd, `COMPASS_PREPARE_OVERWRITE=true`
rebuilds R graphics caches only. Image receipts record the prepared snapshot and DPI;
affected formats regenerate automatically. Existing images without a receipt
are regenerated once.

The notebook's Python kernel requires Polars. R needs its existing plotting
packages plus `nanoparquet` or `arrow`; no Python installation is needed to knit.
The notebook controls `POLARS_MAX_THREADS`, defaulting to at most four and
respecting the SLURM CPU allocation when available.
`COMPASS_PREPARE_WORKERS` defaults to one to limit memory use;
`COMPASS_RENDER_WORKERS` defaults to two and parallelizes individual panels on
Linux. macOS and Windows render sequentially because native graphics libraries
are not safe to fork.

## Cohorts and outputs

Defaults: all ADT, pre-ADT castrate patients included, platinum and NEPC, at
landmarks 0/90/180. The forests read existing day-180 results across six ADT
configurations without enabling six full renders. ARPI uses its own input file.
Retired AVPC endpoints and localized-adjuvant cohorts remain unavailable.

Canonical ADT/platinum panels are main figures; NEPC, cohort variants and
diagnostics are supplements. Individual subpanels retain this layout:

`FIG_ROOT/<ARM>/by_figure/<main|supplements>/<group>/<artifact>/<endpoint>__<subset>__<exclusion>.png`

Figure 2 runs once for canonical ADT/platinum. Lab figures run on each selected
platinum cohort, for PSA/testosterone by default. Trajectories retain day-zero
boundaries, 180-day bins, and log transformation before averaging. Pre-ADT
coverage uses five years and full-stratum denominators. Lab KM curves compare
bottom versus top 20% without splitting ties. Somatic and Gleason curves retain
their specimen/score date origins. GAMs, non-androgen lab panels and the older
ADT-intent comparison are disabled by default.

The metastatic supplement defaults to three heatmaps: ADT/LLM versus maximum
regex stage across the record, and ADT versus LLM. Set
`COMPASS_METASTATIC_EXTRA_PANELS=1` to restore coverage, ICD burden, survival and
trajectory diagnostics. These use canonical ADT/platinum and the longitudinal
analysis anchor, with supplied ADT/LLM labels preserved.

The federated comparison runs when canonical ADT/platinum is selected, or scope
is `federated`. Missing optional sources are reported and skipped. Failures in
one figure set do not prevent preparation/rendering of other available sets.

## Configuration

```{r setup}
options(readr.num_threads = 1L, readr.read_lazy = FALSE)
pipeline_path <- file.path(getwd(), "COMPASS_generate_figures_pipeline.R")
if (!file.exists(pipeline_path)) pipeline_path <- file.path(
  getwd(), "COMPASS", "survival_analysis", "COMPASS_generate_figures_pipeline.R")
pipeline_path <- normalizePath(pipeline_path, mustWork = TRUE)
source(pipeline_path)
source(file.path(dirname(pipeline_path), "figure_data_cache.R"))

env_flag <- function(name, default = FALSE) {
  value <- tolower(Sys.getenv(name, if (default) "true" else "false"))
  if (!value %in% c("true", "false", "1", "0", "yes", "no")) stop(name, " must be true or false")
  value %in% c("true", "1", "yes")
}
env_values <- function(name, default) trimws(strsplit(Sys.getenv(name, default), ",", fixed = TRUE)[[1]])
env_workers <- function(name, default) {
  value <- suppressWarnings(as.integer(Sys.getenv(name, as.character(default))))
  if (is.na(value) || value < 1L) stop(name, " must be a positive integer")
  value
}
NEPC_PROJ_PATH <- Sys.getenv("COMPASS_DATA_ROOT", "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS")
FIG_ROOT <- Sys.getenv("COMPASS_FIG_ROOT", "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS")
FIGURE_DATA_ROOT <- Sys.getenv("COMPASS_FIGURE_DATA_ROOT", file.path(NEPC_PROJ_PATH, "figure_data"))
FIGURE_STAGE <- match.arg(Sys.getenv("COMPASS_FIGURE_STAGE", "all"), c("all", "prepare", "render"))
FIGURE_SCOPE <- match.arg(Sys.getenv("COMPASS_FIGURE_SCOPE", "all"), c("all", "federated"))
ENDPOINTS <- env_values("COMPASS_FIGURE_ENDPOINTS", "nepc,platinum")
ADT_COHORT_OPTIONS <- c("adt", "adt_noprecastrate", "adt_metastatic_adt",
  "adt_metastatic_adt_noprecastrate", "adt_metastatic_llm", "adt_metastatic_llm_noprecastrate")
COHORTS <- env_values("COMPASS_FIGURE_COHORTS", "adt")
stopifnot(length(COHORTS) > 0L, length(ENDPOINTS) > 0L,
          all(COHORTS %in% SUPPORTED_COHORTS), all(ENDPOINTS %in% SUPPORTED_ENDPOINTS),
          !anyDuplicated(COHORTS), !anyDuplicated(ENDPOINTS))
COHORT_FOREST_CONFIG <- list(script = file.path(dirname(pipeline_path), "cohort_forest_figures.R"),
                             cohorts = ADT_COHORT_OPTIONS, landmark = 180L)
PLOT_NON_ANDROGEN_LAB_FIGURES <- env_flag("COMPASS_NON_ANDROGEN_LABS")
PLOT_GAM_TRAJECTORIES <- env_flag("COMPASS_GAM_TRAJECTORIES")
PLOT_ADT_INTENT_SUPPLEMENT <- env_flag("COMPASS_ADT_INTENT_SUPPLEMENT")
PLOT_METASTATIC_LABEL_SUPPLEMENT <- env_flag("COMPASS_METASTATIC_LABEL_SUPPLEMENT", TRUE)
PLOT_FEDERATED_NO_MSK_SUPPLEMENT <- env_flag("COMPASS_FEDERATED_NO_MSK_SUPPLEMENT", TRUE)
METASTATIC_SUPPLEMENT <- list(
  prepare_script = file.path(dirname(pipeline_path), "prepare_metastatic_figure_labels.R"),
  r_script = file.path(dirname(pipeline_path), "metastatic_figure_supplements.R"),
  intent = file.path(NEPC_PROJ_PATH, "mrn_lists", "adt_intent_labels_model_cohort.csv"),
  stage = Sys.getenv("COMPASS_REGEX_STAGE_PATH", file.path(
    Sys.getenv("PROFILE_DATA_PATH", "/data/gusev/USERS/jpconnor/data/PROFILE_DATA"),
    "CANCER_ANNOTATIONS", "CANCER_STAGE_NOTE_LEVEL.parquet")),
  llm = Sys.getenv("LLM_MET_LABELS_PATH", "/data/gusev/USERS/jpconnor/data/LLM_annotations/LLM_met_diagnosis/met_dx_labels.parquet"),
  icd = file.path(NEPC_PROJ_PATH, "prostate_icd_data.csv"))
FEDERATED_NO_MSK_SUPPLEMENT <- list(
  script = file.path(dirname(pipeline_path), "federated_no_msk_figures.R"),
  results = Sys.getenv("COMPASS_FEDERATED_NO_MSK_RESULTS", file.path(
    NEPC_PROJ_PATH, "federated_results_no_MSK", "cox_federated_univariate_adt.csv")))
RENDER_PROFILE <- match.arg(tolower(Sys.getenv("COMPASS_RENDER_PROFILE", "fast")), c("fast", "publication"))
RENDER_DPI <- if (RENDER_PROFILE == "publication") 600 else 200
RENDER_PDF <- env_flag("COMPASS_RENDER_PDF")
RENDER_OVERWRITE <- env_flag("COMPASS_RENDER_OVERWRITE")
PREPARE_OVERWRITE <- env_flag("COMPASS_PREPARE_OVERWRITE")
RENDER_WORKERS <- env_workers("COMPASS_RENDER_WORKERS", 2L)
PREPARE_WORKERS <- env_workers("COMPASS_PREPARE_WORKERS", 1L)
source(file.path(dirname(pipeline_path), "metastatic_figure_supplements.R"))

figure_config <- list(
  data_root = NEPC_PROJ_PATH, cache_root = FIGURE_DATA_ROOT, fig_root = FIG_ROOT,
  cohorts = COHORTS, endpoints = ENDPOINTS, scope = FIGURE_SCOPE,
  labs = if (PLOT_NON_ANDROGEN_LAB_FIGURES) names(CATEGORY_MAP) else ANDROGEN,
  classifier_path = DEFAULT_LLM_ANNOTATIONS_PATH,
  forest_cohorts = COHORT_FOREST_CONFIG$cohorts, forest_landmark = COHORT_FOREST_CONFIG$landmark,
  gam = PLOT_GAM_TRAJECTORIES, adt_intent = PLOT_ADT_INTENT_SUPPLEMENT,
  metastatic = PLOT_METASTATIC_LABEL_SUPPLEMENT, metastatic_extra = metastatic_extra_panels_enabled(),
  metastatic_sources = METASTATIC_SUPPLEMENT[c("intent", "stage", "llm", "icd")],
  federated = PLOT_FEDERATED_NO_MSK_SUPPLEMENT,
  federated_path = FEDERATED_NO_MSK_SUPPLEMENT$results)
```

## Prepare and render

```{r figures}
figure_run <- run_cached_figure_workflow(
  figure_config, pipeline_path, FIGURE_STAGE,
  prepare_workers = PREPARE_WORKERS, render_workers = RENDER_WORKERS,
  dpi = RENDER_DPI, pdf = RENDER_PDF,
  prepare_overwrite = PREPARE_OVERWRITE, render_overwrite = RENDER_OVERWRITE,
  metastatic_config = METASTATIC_SUPPLEMENT,
  forest_config = COHORT_FOREST_CONFIG,
  federated_config = FEDERATED_NO_MSK_SUPPLEMENT)
```
