---
title: "COMPASS platinum figures and supplementary analyses"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Canonical R figure workflow using tidyverse/ggplot2. By default, **Knit**
renders the full **ADT** cohort with pre-ADT castrate patients included for
the **NEPC** and **platinum** endpoints. Metastatic subsets, the pre-ADT
castrate exclusion, and ARPI remain selectable in the configuration below.

The AVPC endpoint and the localized-adjuvant cohort are retired; no such trees
are built. Figure 2 uses the four primary subtypes for its landscape and
aggressive-variant enrichment panels. Lab-trajectory stratifications remain
limited to platinum and binary NEPC status.

The document reads the merged `profile_data` run under `.../CAIA/COMPASS`
and writes to `figures/CAIA/COMPASS`.

Platinum predictions for the full ADT cohort with pre-ADT castrate patients
included are the **main figures**. NEPC results, alternative cohort definitions,
NEPC-label validation, and ADT/LLM/regex diagnostics are **supplements**.
All cohort configurations remain selectable; the default stays `COHORTS <- c("adt")`.

Outputs are grouped by figure and individual subpanel:

`FIG_ROOT/<ARM>/by_figure/<main|supplements>/<group>/<artifact>/<endpoint>__<subset>__<exclusion>.png`

For example:

```
ADT/by_figure/main/figure3/univariate_landmark180/platinum__all__incl.png
ADT/by_figure/supplements/figure3/univariate_landmark180/nepc__all__incl.png
ADT/by_figure/supplements/figure3/univariate_landmark180/platinum__metastatic_adt__incl.png
ADT/by_figure/supplements/metastatic_labels/coverage/platinum__all__incl.png
```

Artifact names omit their parent figure prefix and lab name. Completed panels
from the preceding `by_figure/<group>/<artifact>/` layout are copied to the new
main/supplement folders when the exact cohort/endpoint leaf matches, avoiding
unnecessary rendering. Originals are retained. Overwrite regenerates instead.

Existing figures at the exact
cohort/endpoint output paths are skipped by default. Set
`COMPASS_RENDER_OVERWRITE=true` (or `RENDER_OVERWRITE <- TRUE` below) to regenerate
them after changing inputs, styling, or DPI. Unrelated older outputs are left
alone. Completion checks inspect file headers and end markers; empty or
truncated graphics are regenerated.

PNG is the default required figure format. Set
`COMPASS_RENDER_PDF=true` to additionally write a vector PDF beside every PNG;
tables remain CSV/Markdown and follow the same leaf-naming convention. There is
no `R/` language subdirectory.

When PDF output is enabled, a panel is complete only when both PNG and PDF
exist. Missing formats are generated individually, preserving completed ones.
New graphics are written to temporary files and published only after a
successful save. Input preparation and CSV/Markdown tables still run.

Figures are saved as individual subpanels only. Figure 1 writes its CONSORT,
KM, and three timing panels; Figure 2 writes its confusion matrix, metrics,
subtype landscape, and platinum enrichment;
Figure 4 writes AUC, C-index, and each model/landmark importance panel.
Assembled composite figures are no longer constructed or saved in either
rendering profile.

Figure 2 v3 validates the binary `has_nepc` classifier call over every
ADT-exposed patient. It is emitted once from the canonical `adt` / platinum
pass and skipped for the other cells. Panels A1/A2 show the binary NEPC
confusion matrix and classifier metrics. Restored Panel B shows the
conventional/AVPC/NEPC/biomarker landscape within each platinum-status group.
Restored Panel C compares platinum use among aggressive primary subtypes
(AVPC + NEPC) versus conventional disease, excluding biomarker/unclassified
labels, with 95% Wilson intervals and a one-sided Fisher test. B/C use all
classifier-labeled ADT-exposed patients, independently of manual chart review.
AVPC confusion panels remain retired.

All lab outputs under `by_figure/<main|supplements>/labs/` are likewise emitted only by the platinum
pass. Lab outputs include longitudinal summaries and time-to-platinum tertile curves; R-fitted GAM panels
remain available but are toggled off by default. The distribution and
quartile-KM families remain retired; the restored curves use all three tertiles. A second NEPC-labelled copy would not
represent a different endpoint analysis.
Longitudinal summaries use 180-day bins from one year before through five
years after the treatment anchor, with day 0 as a hard boundary. A second
trajectory set can be fit directly during plotting with `mgcv::bam` when
`PLOT_GAM_TRAJECTORIES` is enabled; fREML and shrinkage select smoothness
without depending on Python GAM outputs. Both implementations support the
original scale and log space. The log versions transform each lab measurement
with `log1p` before patient/bin aggregation or GAM fitting and use an `_log`
filename suffix.
Trajectory and GAM titles contain only the cohort definition and contributing
patient count; the lab, scale, and stratum remain identified by the axes,
legend, and artifact path.
For the ADT arm, two additional PSA/testosterone diagnostics quantify the
pre-ADT data supporting those trajectories: any/recent measurement coverage
and zero-anchored 180-day-bin availability. Platinum panels use the
entire analysis-cohort stratum as their denominator; NEPC panels use all
classifier-labeled cohort patients, including those with no androgen lab. The
coverage diagnostics retain their five-year pre-ADT lookback independently of
the shorter trajectory window.

Figure 3b includes sequencing-based somatic associations (volcano plus top-hits
forest) and a separate Gleason-score forest. Each uses its own index date and
existing Cox results for the selected endpoint. PRS panels remain retired.
These analyses exist for the ADT arm at landmark 0 only; a missing
`cox_somatic_gleason/` result is reported and skipped independently.

The platinum pass now writes PSA and testosterone **tertile KM curves** at
each selected landmark, using the pre-landmark mean in `aggregated_landmark*.csv`.
Only incident, observed outcomes enter the split. Equal lab values stay in one
group; a distribution that cannot form three nonempty value-based groups is
reported and skipped. Legends show each group's patient and event counts.

Figure 3b also writes platinum KM curves for every significant mutation feature
(Cox q < 0.05), comparing carriers with non-carriers among patients with a known
binary call. These curves start at the sequencing specimen date. Gleason KM
curves start at the selected score date, grouping scores as ≤6, 7, 8, 9, and 10.
Both read the existing indexed `somatic_gleason/<analysis>/aggregated_landmark0.csv`
inputs. They do not mix those clocks with ADT-relative durations. Mutation
selection and KM visualization use the same cohort, so the panels are descriptive.

Two additional cohort forests show **PSA/testosterone for platinum** and
**PSA/testosterone for NEPC**. Each has mean, minimum, maximum, and last-value
rows; delta and observation counts are excluded. `COHORT_FOREST_CONFIG` controls
the landmark (180 days by default) and available cohorts. These summaries read
existing results from all six ADT configurations without enabling six full
figure renders. The forest section of `07_cohort_comparison.ipynb` uses the same
`cohort_forest_figures.R` renderer and output paths.

The local ADT figure review prompted wrapped titles/captions, external legends,
clearer background points, fewer overlapping volcano labels, and axes that
retain all plotted estimates. Run once with **`COMPASS_RENDER_OVERWRITE=true`**
to replace existing panels with the revised styling; then return it to `false`
for normal resume behavior. The downloaded PNGs contain no patient-level KM
inputs, so the new survival curves require the corresponding analysis inputs.

`05_figures.Rmd` now also renders the ADT filtering diagnostics as separate
supplemental subpanels. These use the canonical ADT platinum time-zero analysis
cohort, including pre-ADT castrate patients, so counts are comparable across
ADT intent and LLM metastatic status. This differs from the older
`08_adt_intent.ipynb` all-ADT-exposed population.

Regex stage is collapsed to **I–III = local** and **IV = metastatic**. Missing
or invalid stages remain unclassified. The primary comparison uses the nearest
stage within 365 days before/on ADT; maximum stage before/on ADT and maximum
stage after ADT are separate retrospective comparisons. Each compares ADT
intent with regex and LLM metastatic status with regex; an additional panel
compares ADT with LLM. Heatmaps report counts, row percentages, paired coverage,
and agreement. A coverage panel retains missing labels. No method is designated
a gold standard. ADT intent and the LLM metastatic label both use full observed
history, so these are descriptive comparisons rather than baseline validation.

The same supplement includes ICD site burden, site patterns, death/platinum/NEPC
KM curves, and PSA/testosterone trajectories for each label definition, where
inputs are available. It reuses cached longitudinal data, saves only individual
subpanels, and shares progress tracking and the overwrite switch. The older
localized-adjuvant versus metastatic **model-result** comparison remains
available via `PLOT_ADT_INTENT_SUPPLEMENT`; it needs precomputed comparison CSVs.

`PLOT_METASTATIC_LABEL_SUPPLEMENT` defaults to `TRUE`. Preparation runs entirely
in R, using `arrow` or `nanoparquet` for the Parquet inputs and the existing
`mrn_lists/adt_intent_labels_model_cohort.csv`, dated regex stage parquet, and
`LLM_met_diagnosis/met_dx_labels.parquet`. Set
`COMPASS_REGEX_STAGE_PATH` or `LLM_MET_LABELS_PATH` to use other input paths.
If neither R Parquet reader is installed, run `install.packages("nanoparquet")`
once in the R environment used for knitting. No Python interpreter is called.
The NEPC subtype classifier is not used as a metastatic-status label.
Missing required sources produce a warning and skip
this supplement; other figures continue.

The main workflow calls `generate_figures()` in
`COMPASS_generate_figures_pipeline.R`. Supplement rendering lives in
`metastatic_figure_supplements.R`; `prepare_metastatic_figure_labels.R` joins
small label/stage inputs without reading the longitudinal lab file.

The Figure 1 analysis-set supplement, Figure 3 significance-only variants,
Figure 4 all-model supplement, and subtype-stratified KM family are retired.

Run `01_preprocessing.ipynb` to completion first; this document only reads
that merged-profile run's outputs.

## Setup

```{r setup}
# Set these before sourcing/preloading: each fork is already a parallel task.
# readr otherwise may start a parser thread per visible CPU in every worker;
# lazy reads can also defer parsing of supposedly warm inputs until after fork.
options(readr.num_threads = 1L, readr.read_lazy = FALSE)

pipeline_path <- normalizePath(
  file.path(getwd(), "COMPASS_generate_figures_pipeline.R"),
  mustWork = FALSE
)
if (!file.exists(pipeline_path)) {
  pipeline_path <- normalizePath(
    file.path(getwd(), "COMPASS", "survival_analysis", "COMPASS_generate_figures_pipeline.R"),
    mustWork = FALSE
  )
}
if (!file.exists(pipeline_path))
  stop("Could not locate COMPASS_generate_figures_pipeline.R")
source(pipeline_path)

NEPC_PROJ_PATH <- "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS"
FIG_ROOT <- "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS"

# Endpoints to render. Each needs its own 01/02/03 build; see
# compass_pipeline.make_endpoint_runs(). Must be a subset of the pipeline's
# SUPPORTED_ENDPOINTS -- naming a retired endpoint here fails in the
# ENDPOINT_SUFFIXES lookup below rather than at generate_figures().
ENDPOINTS <- c("nepc", "platinum")

# Keep every modelled ADT configuration available. To render them all, set
# COHORTS <- ADT_COHORT_OPTIONS. ARPI labels remain in SUPPORTED_COHORTS.
ADT_COHORT_OPTIONS <- c(
  "adt",                 "adt_noprecastrate",
  "adt_metastatic_adt",  "adt_metastatic_adt_noprecastrate",
  "adt_metastatic_llm",  "adt_metastatic_llm_noprecastrate"
)
COHORTS <- c("adt")  # all ADT patients; pre-ADT castrate patients included
# This small, read-only summary compares available results from all six ADT
# configurations without enabling the expensive full render for each variant.
COHORT_FOREST_CONFIG <- list(
  script = file.path(dirname(pipeline_path), "cohort_forest_figures.R"),
  cohorts = ADT_COHORT_OPTIONS,
  landmark = 180L
)

stopifnot(
  all(ENDPOINTS %in% SUPPORTED_ENDPOINTS),
  all(COHORTS %in% SUPPORTED_COHORTS)
)

# Generate lab-specific panels only for PSA and Testosterone.
# Non-androgen labs remain available to models and aggregate feature plots.
PLOT_NON_ANDROGEN_DISTRIBUTIONS <- FALSE
PLOT_NON_ANDROGEN_LAB_FIGURES <- FALSE
# Keep the R/mgcv fitting implementation available without running its
# comparatively expensive trajectory fits during the default figure knit.
PLOT_GAM_TRAJECTORIES <- FALSE
# The ADT-intent comparison supplement runs after Figure 4 in the canonical ADT
# cell. Keep its implementation available without extending the primary render.
PLOT_ADT_INTENT_SUPPLEMENT <- FALSE

# Merge the ADT filtering diagnostics into this workflow. Uses the medication
# intent labels already built by compass_pipeline; no model refits are needed.
PLOT_METASTATIC_LABEL_SUPPLEMENT <- TRUE
METASTATIC_SUPPLEMENT <- list(
  prepare_script = file.path(dirname(pipeline_path), "prepare_metastatic_figure_labels.R"),
  r_script = file.path(dirname(pipeline_path), "metastatic_figure_supplements.R"),
  intent = file.path(NEPC_PROJ_PATH, "mrn_lists", "adt_intent_labels_model_cohort.csv"),
  stage = Sys.getenv("COMPASS_REGEX_STAGE_PATH", file.path(
    Sys.getenv("PROFILE_DATA_PATH", "/data/gusev/USERS/jpconnor/data/PROFILE_DATA"),
    "CANCER_ANNOTATIONS", "CANCER_STAGE_NOTE_LEVEL.parquet")),
  llm = Sys.getenv("LLM_MET_LABELS_PATH",
    "/data/gusev/USERS/jpconnor/data/LLM_annotations/LLM_met_diagnosis/met_dx_labels.parquet"),
  icd = file.path(NEPC_PROJ_PATH, "prostate_icd_data.csv")
)

# Rendering profile. Both profiles save individual subpanels only, at 200 dpi
# by default. Use COMPASS_RENDER_PROFILE=publication for 600-dpi subpanels and
# enable overwrite to replace figures already rendered at a different DPI.
RENDER_PROFILE <- tolower(Sys.getenv("COMPASS_RENDER_PROFILE", "fast"))
if (!RENDER_PROFILE %in% c("fast", "publication"))
  stop("COMPASS_RENDER_PROFILE must be 'fast' or 'publication'")
RENDER_DPI <- if (RENDER_PROFILE == "publication") 600 else 200
RENDER_OUTPUT_MODE <- "panels"

pdf_value <- tolower(Sys.getenv("COMPASS_RENDER_PDF", "false"))
if (!pdf_value %in% c("true", "false", "1", "0", "yes", "no"))
  stop("COMPASS_RENDER_PDF must be true or false")
RENDER_PDF <- pdf_value %in% c("true", "1", "yes")

# FALSE resumes missing figures; TRUE regenerates all requested figure files.
overwrite_value <- tolower(Sys.getenv("COMPASS_RENDER_OVERWRITE", "false"))
if (!overwrite_value %in% c("true", "false", "1", "0", "yes", "no"))
  stop("COMPASS_RENDER_OVERWRITE must be true or false")
RENDER_OVERWRITE <- overwrite_value %in% c("true", "1", "yes")

# Parallelize cohort x endpoint cells, but keep the default conservative: each
# render holds plotting data, so launching all twelve cells
# together can exhaust memory and make the knit appear to freeze. Increase the
# worker count explicitly only on a host with enough RAM.
RENDER_WORKERS <- suppressWarnings(as.integer(
  Sys.getenv("COMPASS_RENDER_WORKERS", "2")
))
if (is.na(RENDER_WORKERS) || RENDER_WORKERS < 1L)
  stop("COMPASS_RENDER_WORKERS must be a positive integer")

# Check the R Parquet reader before expensive reads/forking, only when this
# supplement needs rendering. Completed figures do not need source readers.
if (PLOT_METASTATIC_LABEL_SUPPLEMENT && "adt" %in% COHORTS && "platinum" %in% ENDPOINTS &&
    all(file.exists(unlist(METASTATIC_SUPPLEMENT[c("intent", "stage", "llm")])))) {
  source(METASTATIC_SUPPLEMENT$r_script)
  bases <- file.path(FIG_ROOT, "ADT", "by_figure", "supplements", "metastatic_labels",
                     sub("^adt_labels_", "", metastatic_supplement_stems()), "platinum__all__incl")
  outputs <- c(paste0(bases, ".png"), if (RENDER_PDF) paste0(bases, ".pdf"))
  if (RENDER_OVERWRITE || !all(vapply(outputs, figure_file_complete, logical(1)))) {
    source(METASTATIC_SUPPLEMENT$prepare_script)
    metastatic_parquet_backend()
  }
}

# Fail early with a useful message rather than deep inside generate_figures().
for (required_path in c(
    file.path(NEPC_PROJ_PATH, "LLM_NEPC_labels"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "platinum_MRN_list.csv"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "icd_prostate_mrn_flags.csv"),
    file.path(NEPC_PROJ_PATH, "longitudinal_prediction_data_adt.csv")
)) {
    if (!file.exists(required_path))
      stop(required_path, " not found -- run 01_preprocessing.ipynb first.")
}

# Each cohort x endpoint reads its own prediction-inputs tree. Check them all up
# front and report EVERY missing one together: with 6 cohorts x 2 endpoints,
# stopping at the first would mean 12 edit-rerun cycles to discover an
# incomplete 01_preprocessing pass.
missing_trees <- character(0)
for (COHORT in COHORTS) {
  for (ENDPOINT in ENDPOINTS) {
    suffix <- unname(ENDPOINT_SUFFIXES[[ENDPOINT]])
    inputs_dir <- file.path(NEPC_PROJ_PATH, "survival_analysis",
                            paste0("prediction_inputs_", COHORT, suffix))
    if (!dir.exists(inputs_dir))
      missing_trees <- c(missing_trees, inputs_dir)
  }
}
if (length(missing_trees))
  stop(length(missing_trees), " of ", length(COHORTS) * length(ENDPOINTS),
       " prediction-input trees are missing -- run 01_preprocessing.ipynb for ",
       "every cohort x endpoint first:\n  ",
       paste(missing_trees, collapse = "\n  "))

# Populate arm-level caches before parallel::mclapply forks. On Unix, workers
# then inherit these large read-only objects through copy-on-write memory
# instead of parsing the longitudinal CSV independently up to 12 times.
# Sequential/Windows renders benefit from the same warm caches.
shared_longitudinal_csv <- file.path(
  NEPC_PROJ_PATH, "longitudinal_prediction_data_adt.csv"
)
preload_started <- proc.time()[["elapsed"]]
message("Preloading patient metadata and all-lab record spans ...")
invisible(cached_profile_patient_and_labs(shared_longitudinal_csv, "DFCI_MRN"))
message(sprintf("Patient metadata and record spans ready after %.1f seconds; preloading trajectories ...",
                proc.time()[["elapsed"]] - preload_started))
invisible(cached_canonical_longitudinal(
  shared_longitudinal_csv,
  labs = if (PLOT_NON_ANDROGEN_LAB_FIGURES) names(CATEGORY_MAP) else ANDROGEN
))

# The wide longitudinal CSV is no longer needed once both processed caches
# are ready. Drop that raw frame before warming the smaller shared inputs.
clear_raw_read_cache()

# These sources are smaller, but every cell reads them. Warm their raw caches
# as well so forked workers do no repeated shared-input I/O.
invisible(cached_read_csv(
  file.path(NEPC_PROJ_PATH, "LLM_NEPC_labels", "baca_lab_annotations.csv"),
  show_col_types = FALSE
))
invisible(cached_read_csv(
  file.path(NEPC_PROJ_PATH, "mrn_lists", "platinum_MRN_list.csv"),
  show_col_types = FALSE
))
invisible(cached_read_csv(
  file.path(NEPC_PROJ_PATH, "mrn_lists", "icd_prostate_mrn_flags.csv"),
  show_col_types = FALSE
))
invisible(load_llm_strata(DEFAULT_LLM_ANNOTATIONS_PATH))
# Collect temporary parsing/transformation objects before workers inherit the
# heap. Cached patient summaries and compact trajectory tables remain live.
invisible(gc())
message(sprintf(
  "shared inputs preloaded once in %.1f seconds; forking workers from warm cache",
  proc.time()[["elapsed"]] - preload_started
))

message("data root:   ", NEPC_PROJ_PATH)
message("figure root: ", FIG_ROOT)
message("endpoints:   ", paste(ENDPOINTS, collapse = ", "))
message("cohorts:     ", paste(COHORTS, collapse = ", "))
message(sprintf("renders:     %d cohort x endpoint figure sets",
                length(COHORTS) * length(ENDPOINTS)))
message(sprintf("profile:     %s (%d dpi, output=%s, pdf=%s, overwrite=%s, workers=%d)",
                RENDER_PROFILE, RENDER_DPI, RENDER_OUTPUT_MODE,
                RENDER_PDF, RENDER_OVERWRITE, RENDER_WORKERS))
```

## Render ADT figures

Runs the R figure-generation pipeline once per selected cohort x endpoint.
The default is 1 cohort x 2 endpoints = 2 figure sets. Outputs go to
`FIG_ROOT/<ARM>/by_figure/<main|supplements>/`, with
each run writing one `<endpoint>__<subset>__<exclusion>` leaf per plot stem.

Set `COHORTS <- ADT_COHORT_OPTIONS` to render all six ADT configurations,
or select individual cohort labels and endpoints above. `COHORTS`
accepts any run label in `SUPPORTED_COHORTS`, so the ARPI arm is rendered by
naming its supported labels. Retired cohort and endpoint labels are rejected.

Each cohort x endpoint cell is rendered independently, with two cells running
concurrently by default. A failure in one is reported rather than abandoning
the remaining cells. Any failures are re-reported at the end and the chunk then fails,
so a partial render cannot pass silently.

Shared inputs are parsed eagerly with one reader thread. Patient record spans
are computed once per arm, and only requested labs enter the trajectory cache.
The log reports preload time, each PNG/PDF render time, and total time per cell,
so a slow input transformation can be distinguished from a slow graphics device.

A live console progress bar combines both workers: finished cohort/endpoint
sets out of the total, subpanels saved or skipped, failures, elapsed time, and the worker's
current figure or preparation stage. The percentage tracks sets, since optional
input availability changes the number of subpanels. A PNG and its optional PDF
count as one completed subpanel after both requested files are available.

```{r render-figures}
render_cell <- function(COHORT, ENDPOINT, cell_index) {
  label <- sprintf("%s / %s", cohort_display(COHORT), toupper(ENDPOINT))
  cell_started <- proc.time()[["elapsed"]]
  current_stage <- "initialization"
  progress <- function(event, detail = "") {
    if (event == "stage" && nzchar(detail)) current_stage <<- detail
    figure_progress$update(cell_index, event, detail)
    invisible(NULL)
  }
  progress("start")
  failure_detail <- ""
  message(sprintf("\n========== R figures: %s ==========", label))
  ok <- tryCatch({
    generate_figures(
      COHORT,
      nepc_proj_path = NEPC_PROJ_PATH,
      fig_root = FIG_ROOT,
      endpoint = ENDPOINT,
      plot_non_androgen_distributions = PLOT_NON_ANDROGEN_DISTRIBUTIONS,
      plot_non_androgen_lab_figures = PLOT_NON_ANDROGEN_LAB_FIGURES,
      plot_gam_trajectories = PLOT_GAM_TRAJECTORIES,
      plot_adt_intent_supplement = PLOT_ADT_INTENT_SUPPLEMENT,
      save_dpi = RENDER_DPI,
      save_pdf = RENDER_PDF,
      output_mode = RENDER_OUTPUT_MODE,
      progress = progress,
      overwrite = RENDER_OVERWRITE,
      metastatic_supplement = if (PLOT_METASTATIC_LABEL_SUPPLEMENT) METASTATIC_SUPPLEMENT else NULL,
      cohort_forest_config = COHORT_FOREST_CONFIG
    )
    TRUE
  }, error = function(e) {
    failure_detail <<- sprintf("%s [%s]: %s", label, current_stage, conditionMessage(e))
    failed_call <- conditionCall(e)
    if (!is.null(failed_call))
      failure_detail <<- paste0(failure_detail, "\n    Call: ",
                               paste(deparse(failed_call), collapse = " "))
    message("FAILED ", failure_detail)
    FALSE
  })
  if (ok) {
    progress("complete")
    message(sprintf("\nCompleted R figures for %s in %.1f seconds.", label,
                    proc.time()[["elapsed"]] - cell_started))
    character(0)
  } else {
    progress("failed", failure_detail)
    failure_detail
  }
}

render_grid <- tidyr::crossing(COHORT = COHORTS, ENDPOINT = ENDPOINTS)
figure_progress <- new_figure_progress(
  paste(render_grid$COHORT, toupper(render_grid$ENDPOINT), sep = " / ")
)
render_one <- function(i) {
  render_cell(render_grid$COHORT[[i]], render_grid$ENDPOINT[[i]], i)
}
n_workers <- min(RENDER_WORKERS, nrow(render_grid))
render_results <- tryCatch({
if (n_workers > 1L && .Platform$OS.type != "windows") {
  # Platinum cells emit the additional longitudinal panels and are much slower
  # than NEPC cells. Dynamic scheduling prevents one prescheduled fork from
  # receiving every heavy platinum cell while the NEPC fork finishes early.
  parallel::mclapply(
    seq_len(nrow(render_grid)), render_one,
    mc.cores = n_workers, mc.preschedule = FALSE
  )
} else {
  if (n_workers > 1L)
    message("Parallel fork rendering is unavailable on Windows; using one worker.")
  lapply(seq_len(nrow(render_grid)), render_one)
}
}, error = function(e) {
  figure_progress$close()
  stop(e)
}, interrupt = function(e) {
  figure_progress$close()
  stop(e)
})
worker_errors <- vapply(render_results, function(result) {
  # mclapply can return NULL when a worker is killed (for example by OOM).
  # A successful cell returns character(0), which must stay distinct from NULL.
  is.null(result) || inherits(result, "try-error")
}, logical(1))
for (i in which(worker_errors))
  figure_progress$update(i, "failed", "worker exited without a result")
render_failures <- unlist(render_results[!worker_errors], use.names = FALSE)
if (any(worker_errors))
  render_failures <- c(
    render_failures,
    unlist(lapply(which(worker_errors), function(i) {
      sprintf("%s / %s (worker failed)", cohort_display(render_grid$COHORT[[i]]),
              toupper(render_grid$ENDPOINT[[i]]))
    }), use.names = FALSE)
  )

n_total <- length(COHORTS) * length(ENDPOINTS)
progress_summary <- figure_progress$snapshot()
figure_progress$close()
message(sprintf("%d subpanels saved, %d completed subpanels skipped in %.1f minutes.",
                progress_summary$panels, progress_summary$skipped,
                progress_summary$elapsed / 60))
message(sprintf("\n%d of %d figure sets rendered.",
                n_total - length(render_failures), n_total))
if (length(render_failures))
  stop(length(render_failures), " of ", n_total, " figure sets failed:\n  ",
       paste(render_failures, collapse = "\n  "))
```
