---
title: "Generate COMPASS figures (NEPC, AVPC, AVPC/NEPC, and platinum endpoints)"
output:
  html_document:
    toc: true
    toc_depth: 2
---

Canonical R figure workflow using tidyverse/ggplot2. **Knit** renders the
**ADT** cohort for the **NEPC**, **AVPC**, **AVPC/NEPC**, and **platinum**
endpoints by default. ARPI remains available by explicit request.

The document reads the merged `profile_data` run under `.../CAIA/COMPASS`
and writes to `figures/CAIA/COMPASS`.

Outputs are split by cohort arm, then by endpoint, then flattened within each
figure directory. Every endpoint nests one level below its cohort arm --
platinum included, so no endpoint's panels can overwrite another's:

`FIG_ROOT/ADT/nepc/<figure>/adt_<plot-stem>.png`
`FIG_ROOT/ADT/avpc/<figure>/adt_<plot-stem>.png`
`FIG_ROOT/ADT/avpc_nepc/<figure>/adt_<plot-stem>.png`
`FIG_ROOT/ADT/platinum/<figure>/adt_<plot-stem>.png`

Note that platinum figures moved out of the old un-suffixed `FIG_ROOT/ADT/`
path; references to that location need updating. For example,
`figure1/adt_figure1a_consort.png` is written directly in `platinum/figure1/`.
Plot PDFs are not generated. Tables remain CSV/Markdown and use the same flat
figure/cohort-prefix convention. There is no `R/` language subdirectory.

Figure 2 v3 (LLM classifier validation and platinum enrichment over every
ADT-exposed patient) describes the label set and the platinum MRN list rather
than a modelled endpoint, so it is emitted once from the platinum pass and
skipped for the other endpoints. It is the only Figure 2 variant retained;
the earlier v0/v1/v2 variants have been removed.

Figure 3b adds the somatic, Gleason, and PRS univariate associations as their
own plots, one per index analysis. Each is built on its own cohort and index
date (sequencing = specimen collection, Gleason = the score date nearest ADT
start, PRS = ADT start), so they are not merged into the Figure 3 lab
volcanoes. Gleason is drawn as a forest plot because it contributes a single
continuous feature; sequencing and PRS get a volcano plus a top-hits forest.
These analyses exist for the ADT arm at landmark 0 only, and a missing
`cox_somatic_gleason/` tree is reported and skipped.

All figure-generation logic lives in `COMPASS_generate_figures_pipeline.R` as
`generate_figures(cohort, endpoint, ...)`. This document just sources that file
and calls it for each cohort in `COHORTS` (defaults to `c("adt")`) crossed with
each endpoint in `ENDPOINTS` (defaults to all four).

Figure 4 also includes a supplemental held-out comparison of elastic-net Cox,
XGBoost, and Dynamic-DeepHit across mean AUC(t), C-index, and integrated Brier
score. Missing `03b_multivariate_longitudinal.ipynb` outputs are reported in
the figure and its data CSV rather than suppressing the supplement.

Run `01_preprocessing.ipynb` to completion first; this document only reads
that merged-profile run's outputs.

## Setup

```{r setup}
pipeline_path <- normalizePath(
  file.path(getwd(), "COMPASS_generate_figures_pipeline.R"),
  mustWork = FALSE
)
if (!file.exists(pipeline_path)) {
  pipeline_path <- normalizePath(
    file.path(getwd(), "COMPASS", "survival_analysis", "COMPASS_generate_figures_pipeline.R"),
    mustWork = FALSE
  )
}
if (!file.exists(pipeline_path))
  stop("Could not locate COMPASS_generate_figures_pipeline.R")
source(pipeline_path)

NEPC_PROJ_PATH <- "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS"
FIG_ROOT <- "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS"

# Endpoints to render. Each needs its own 01/02/03 build; see
# compass_pipeline.make_endpoint_runs().
ENDPOINTS <- c("nepc", "avpc", "avpc_nepc", "platinum")

# Generate lab-specific panels only for PSA and Testosterone.
# Non-androgen labs remain available to models and aggregate feature plots.
PLOT_NON_ANDROGEN_DISTRIBUTIONS <- FALSE
PLOT_NON_ANDROGEN_LAB_FIGURES <- FALSE

# Fail early with a useful message rather than deep inside generate_figures().
for (required_path in c(
    file.path(NEPC_PROJ_PATH, "LLM_NEPC_labels"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "platinum_MRN_list.csv"),
    file.path(NEPC_PROJ_PATH, "mrn_lists", "icd_prostate_mrn_flags.csv"),
    file.path(NEPC_PROJ_PATH, "longitudinal_prediction_data_adt.csv")
)) {
    if (!file.exists(required_path))
      stop(required_path, " not found -- run 01_preprocessing.ipynb first.")
}

# Each endpoint reads its own prediction-inputs tree. Check them up front so a
# missing NEPC build fails here rather than partway through the platinum pass.
for (COHORT in COHORTS) {
  for (ENDPOINT in ENDPOINTS) {
    suffix <- unname(ENDPOINT_SUFFIXES[[ENDPOINT]])
    inputs_dir <- file.path(NEPC_PROJ_PATH, "survival_analysis",
                            paste0("prediction_inputs_", COHORT, suffix))
    if (!dir.exists(inputs_dir))
      stop(inputs_dir, " not found -- run 01_preprocessing.ipynb for endpoint ",
           ENDPOINT, " first.")
  }
}

message("data root:   ", NEPC_PROJ_PATH)
message("figure root: ", FIG_ROOT)
message("endpoints:   ", paste(ENDPOINTS, collapse = ", "))
```

## Render ADT figures

Runs the R figure-generation pipeline for ADT by default, once per endpoint.
Each endpoint's outputs go to `FIG_ROOT/ADT/<endpoint>/`, i.e. `ADT/nepc/`,
`ADT/avpc/`, `ADT/avpc_nepc/`, and `ADT/platinum/`.
Set `COHORTS <- c("arpi", "adt")` to opt into both arms, or narrow `ENDPOINTS`
to render a single endpoint.

```{r render-figures}
for (COHORT in COHORTS) {
  for (ENDPOINT in ENDPOINTS) {
    message(sprintf("\n========== R figures: %s / %s ==========",
                    COHORT_LABELS[[COHORT]], toupper(ENDPOINT)))
    generate_figures(
      COHORT,
      nepc_proj_path = NEPC_PROJ_PATH,
      fig_root = FIG_ROOT,
      endpoint = ENDPOINT,
      plot_non_androgen_distributions = PLOT_NON_ANDROGEN_DISTRIBUTIONS,
      plot_non_androgen_lab_figures = PLOT_NON_ANDROGEN_LAB_FIGURES
    )
    endpoint_fig_root <- file.path(FIG_ROOT, toupper(COHORT), ENDPOINT)
    supplement_path <- file.path(
      endpoint_fig_root, "figure4",
      paste0(COHORT, "_figure4s_multivariate_all_models.png")
    )
    if (!file.exists(supplement_path))
      stop("Expected multivariate supplement was not generated: ", supplement_path)
    message("multivariate supplement: ", supplement_path)
    message(sprintf("\nCompleted R figures for %s / %s.",
                    COHORT_LABELS[[COHORT]], toupper(ENDPOINT)))
  }
}
```
