---
title: "COMPASS figures: cached preparation and rendering"
output:
  html_document:
    toc: true
    toc_depth: 2
---

## Workflow

Run `04_prep_figure_data.ipynb` first to prepare the Polars tables and manifest.
All Python execution belongs to that notebook. This document and its R helpers
never invoke Python: they consume the notebook's prepared data only. Missing,
changed, or incompatible inputs stop with instructions to rerun the notebook.

Preparation and rendering are separate stages:

1. `04_prep_figure_data.ipynb` uses Polars to scan each selected arm's longitudinal
   CSV once and cache patient metadata, all-lab record spans, selected lab
   measurements, patient/bin raw and log means, and pre-treatment coverage.
2. R consumes those tables, calculates the existing survival statistics and
   optional GAMs, and caches finished graphics objects. CSV/Markdown figure
   tables are written during this stage; no images are generated.
3. Rendering loads the graphics objects and writes individual and multipanel PNGs and optional
   PDFs. It does not read patient inputs, fit curves, or repeat plot calculations.

`COMPASS_FIGURE_STAGE=all` (default) checks the notebook manifest and source
metadata, prepares changed R figure sets, then renders missing or stale formats.
`prepare` builds R graphics caches only; it does not prepare Python tables.
`render` uses the last prepared snapshot without checking
source data, and requires the same cohort/analysis settings. Use `all` for
normal updates and `render` for DPI/PDF changes.

`COMPASS_FIGURE_SCOPE=federated` processes only the no-MSK forests, site cohort report,
and XGBoost performance/importance figures without
requiring longitudinal data or running other figure families.

The site report reads `nvflare_within_site_cox_univariate/cox_within_site_all_sites_cohort.csv`
beside the federated results CSV. It exports cohort sizes, event counts, and observed
event percentages at landmark 0 only in `federated_no_msk/site_incidence_lm000__platinum.{png,csv}`.
The figure matches the cohort-comparison layout: incidence with 95% Wilson CIs
on the left and patient counts with events overlaid on the right. Cells with
fewer than 25 events are hatched/marked. Percentages are events divided by analyzed patients over available follow-up,
not fixed-horizon cumulative incidence estimates. Missing site counts are reported
as unavailable; forest plots still render if the site file is absent.

Caches live under `COMPASS_FIGURE_DATA_ROOT` (default `COMPASS/figure_data`).
They include patient data and need the same access restrictions as the analysis
inputs. All RDS files, including per-image render receipts, live in that data
cache; none are written beside figure exports. On the next knit, recognized
legacy `*.cache.rds` receipts in the figure tree are copied, verified, and moved
into `COMPASS_FIGURE_DATA_ROOT/legacy_render_receipts`.
The data-cache and figure-output directories must be separate and non-nested.
Versioned caches are retained. Source identity uses absolute paths,
file sizes and nanosecond modification times, plus code/settings versions.
Use `FORCE = True` in `04_prep_figure_data.ipynb` after replacing sources while
preserving timestamps and sizes. In this Rmd, `COMPASS_PREPARE_OVERWRITE=true`
rebuilds R graphics caches only. Image receipts record the prepared snapshot and DPI;
affected formats regenerate automatically. Existing images without a receipt
are regenerated once.

The notebook's Python kernel requires Polars. R needs its existing plotting
packages plus `nanoparquet` or `arrow`; no Python installation is needed to knit.
The notebook controls `POLARS_MAX_THREADS`, defaulting to at most four and
respecting the SLURM CPU allocation when available.
`COMPASS_PREPARE_WORKERS` defaults to one to limit memory use;
`COMPASS_RENDER_WORKERS` defaults to two and parallelizes individual panels on
Linux. macOS and Windows render sequentially because native graphics libraries
are not safe to fork.

## Cohorts and outputs

Defaults: all ADT, pre-ADT castrate patients included, platinum only, at
landmarks 0/90/180. NEPC landmark figures are off by default; set
`COMPASS_FIGURE_ENDPOINTS=nepc,platinum` to render both. The forests read
existing day-180 results across six ADT configurations without enabling six
full renders. ARPI uses its own input file.
Retired AVPC endpoints and localized-adjuvant cohorts remain unavailable.

Canonical ADT/platinum also generates two combined cohort-comparison figures:
event incidence with 95% Wilson intervals plus patient/event counts across all
six ADT definitions and both endpoints at day 180; and Stage-1 ADT-intent/LLM
label overlap plus the four agreement-cell counts. The incidence denominator
is each endpoint's aggregated landmark input, not a feature-specific Cox fit.
It is the observed follow-up event fraction, not cumulative risk by day 180.
Cells with fewer than 25 events are hatched/marked; missing cells stay blank.
The label figure uses `mrn_lists/adt_intent_labels_model_cohort.csv` and
`llm_met_labels_model_cohort.csv`, preserving missing labels explicitly.
Run notebook `04` once after this update to prepare these small summary tables.
Combined outputs are under `compiled/cohort/` (count tables remain in `cohort/`):
`event_incidence_lm180__platinum__all__incl` and `metastatic_label_overlap__platinum__all__incl`.

Figure 1 also exports `compiled/cohort/figure1_cohort__<endpoint>__<subset>__<exclusion>`:
a five-panel figure with the cohort-selection flow spanning the left column,
and survival, record span, diagnosis-to-treatment timing, and time to platinum
on the right. Figure 2v3 exports `compiled/labels/figure2v3_llm__platinum__all__incl`,
with agreement, classification metrics, subtype composition, and platinum
enrichment in a 2-by-2 layout. The existing `classifier_validation` and
`subtype_platinum` paired exports remain available. The three metastatic-label
confusion matrices also form `compiled/labels/metastatic_label_agreement__<endpoint>__<subset>__<exclusion>`.
These combined figures have lettered panels and retain each panel's counts,
captions, and legends. All individual panels are exported alongside them.
Compilation stays within each arm/cohort/endpoint; Figure 2v3 remains exclusive
to canonical ADT/platinum. Run the default `COMPASS_FIGURE_STAGE=all` to rebuild
the graphics caches and export these additions; `render` requires a matching prepared snapshot.

The seven-panel DFCI overview is
`compiled/associations/dfci_labs_overview_km_lm0__platinum__<subset>__<exclusion>`.
The top row contains day-0/90/180 within-DFCI univariate volcanoes with common axes;
the second row pairs the PSA log1p trajectory by platinum status with the PSA
bottom-versus-top-20% KM curve; the third row does the same for testosterone.
KM panels default to day 0. Set `COMPASS_COMBINED_KM_LANDMARK=90` or `180` to
select another landmark; the filename and cached graphics record that choice.
The overview requires all seven source panels and is restricted to platinum.
Individual KM panels and the three-landmark KM sheets are also retained.

Exports use shallow topic folders: `cohort/`, `labels/`, `associations/`,
`prediction/`, `trajectories/<lab>/`, and `federated_no_msk/`. There is no
main/supplement split or per-plot directory. Local filenames are
`<figure>__<endpoint>__<subset>__<exclusion>.png`; CSV/Markdown tables live alongside.
Compiled figures have their own `compiled/<topic>/` tree under each arm,
including `compiled/trajectories/<lab>/` to keep PSA/testosterone filenames distinct.
They retain panel letters, with no descriptive panel titles or overall supertitle. Counts formerly in a
supertitle remain in panel subtitles or footnotes. Standalone panels stay in the
original topic folders. Successfully replaced compiled exports from the old
locations are archived in the data cache.

The canonical ADT/platinum run also creates `ADT/manuscript figures/`. These
additional exports use a 7.2-inch manuscript width, consistent 7–9 pt plot text,
lettered panels without subplot titles, and separate `.md` figure legends. They always include a
600-dpi (or higher requested DPI) PNG and a PDF, even when `COMPASS_RENDER_PDF`
is false. The existing individual and compiled exports are retained.

The numbered manuscript set is: `01_cohort_overview`, `02_llm_nepc_avpc`,
`03_dfci_univariable`, `04_multivariable_labs` (held-out Elastic-Net/XGBoost
discrimination versus age baselines), `05_gleason_sensitivity`,
`06_cohort_sensitivity_incidence` (platinum only),
`07_cohort_sensitivity_associations` (day 180),
`08_federated_psa_testosterone`, and `09_federated_xgboost` (test performance
and feature importance). Federated manuscript plots reuse the original subplot
renderers: forests retain site colors and landmark/statistic facets; XGBoost
retains gold model/baseline bars and lab-category importance colors. Incomplete source groups are not assembled. The DFCI
overview uses the same configurable extreme-quintile KM landmark as its larger
compiled counterpart. Figure titles remain in filenames/legends, not supertitles.

Manuscript figures are built from the source ggplots or exported aggregate
tables during `prepare`/`all`, then cached as graphics only. Run `all` after
updating the R files; `render` alone cannot rebuild stale scenes. Cohort counts,
original panel captions and supplied federated population/model-input audit
notes are retained in the accompanying legends. In particular, nonzero gain
for an identifier feature remains an explicit warning on the federated
XGBoost figure and requires a model-input audit before publication. Formatting
does not resolve that issue. Downloaded PNGs alone cannot provide vector
replacements for the original source plots.
Each arm gets a thumbnail `index.html` and a `manifest.csv` with relative paths,
titles, endpoint, cohort, and landmark metadata. No duplicate thumbnail files are made.

R compiles classifier validation into one figure and subtype/enrichment into one;
AUC/C-index into paired comparisons; importance into one three-landmark sheet per
model; lab-extreme KM into one three-landmark sheet per lab; and mutation KM into
pages of at most four panels. Model fits, counts, time origins, and tests are unchanged.
Local univariate lab volcanoes retain individual day-0/90/180 plots and also
export `compiled/associations/labs_univariate_all_landmarks__<endpoint>__<subset>__<exclusion>`.
The combined figure uses three side-by-side panels on a 24 × 9-inch canvas,
common axes, a shared category legend, and separate per-landmark significance
counts. Larger feature labels (approximately 12 pt) and axis text are preserved
in both formats. The cohort and label figures described above also retain their
individual panels, as do the lab KM curves; other compiled figure families do not export component images.

For an existing canonical ADT/platinum PNG download without the graphics cache,
`reformat_downloaded_figures.R <input-ADT-dir> <new-parent>/ADT [0|90|180]` removes
the compiled supertitles and assembles the new overview from the rendered panels.
It preserves raster data/geometry, writes a separate tree and refreshed catalog,
and exports the overview as PNG plus a PDF containing the existing raster panels.
Normal pipeline runs rebuild figures from the underlying R plots instead.

After a successful render, superseded `by_figure` exports for regenerated cohorts
are copied, checksum-verified, and archived in `COMPASS_FIGURE_DATA_ROOT/previous_exports/`;
only then are those old copies and emptied folders removed. Unselected cohorts remain
untouched. Failed renders do not trigger archival. Archives are recoverable in data.

Figure 2 runs once for canonical ADT/platinum. Lab figures run on each selected
platinum cohort, for PSA/testosterone by default. Trajectories retain day-zero
boundaries, 180-day bins, and log transformation before averaging. Pre-ADT
coverage uses five years and full-stratum denominators. Lab KM curves compare
bottom versus top 20% without splitting ties. Somatic and Gleason curves retain
their specimen/score date origins. GAMs, non-androgen lab panels and the older
ADT-intent comparison are disabled by default.

The metastatic supplement defaults to three heatmaps: ADT/LLM versus maximum
regex stage across the record, and ADT versus LLM. Set
`COMPASS_METASTATIC_EXTRA_PANELS=1` to restore coverage, ICD burden, survival and
trajectory diagnostics. These use canonical ADT/platinum and the longitudinal
analysis anchor, with supplied ADT/LLM labels preserved.

The federated HR output contains only PSA/testosterone forest plots at days
0/90/180: mean, minimum, maximum, and last value, with delta and observation-count
features excluded. Two compiled forests (one per analyte) show all sites and the
federated result together, with statistic columns and landmark rows on a 16:9
landscape canvas (one analyte per PowerPoint slide). Plots show
HR per SD and 95% CIs on a common log axis per analyte. Exact p/q values and
modeled/observed counts remain in the adjacent CSV. Symbols distinguish nominal-only p < 0.05,
FDR q < 0.05, and nonsignificance. q-values are not recalculated on this subset.
Finite positive HR estimates are retained; missing/invalid estimates or CIs are
explicitly marked instead of applying the former volcano magnitude filters.
Both across-site and individual within-site HRs are plotted, without re-pooling
site estimates. Within-site forests read
`nvflare_within_site_cox_univariate/cox_within_site_all_sites_results.csv` beside
the across-site results CSV. Each uses its own supplied p/q values.
No matching local results are required. Outputs are
`federated_no_msk/{psa,testosterone}_forest__platinum.{png,csv}` plus the day-0
incidence figure/table. Supplied p/q values are not recomputed or pooled.
The federated population is not assumed to include every displayed site: matching
patient/event counts across landmarks are reported as inferred membership only;
ambiguous or unavailable membership is explicitly unverified.
These forests run when canonical ADT/platinum is selected, or scope
is `federated`. Missing optional sources are reported and skipped. Failures in
one figure set do not prevent preparation/rendering of other available sets.

Federated XGBoost uses `federated_xgboost/xgboost_federated_metrics_adt.csv` and
`xgboost_federated_importance_adt.csv` beside the Cox results. Two additional
compiled figures share the local PROFILE Figure 4 plotting functions:
`federated_no_msk/xgboost_performance__platinum` pairs held-out test mean AUC(t)
and C-index for labs versus the age baseline at days 0/90/180;
`xgboost_importance__platinum` combines the three landmarks' top 15 positive-gain
features, using the same labels, clinical-category colors, and age/body-height
display exclusions as local runs. Unlike the selected HR forests, importance
includes all supplied feature statistics, including deltas and observation counts.
No models are refitted; training/tuning-CV metrics are not substituted for test
performance. The horizon export lacks a train/test identifier and is not plotted.
CSV sidecars retain source metadata, counts, all feature gains, and display flags.
Identifier features with nonzero gain trigger a warning and a figure caption;
they are not silently removed or treated as validated predictors.
Missing optional XGBoost files are reported and skipped independently. Run `04`
again after this update to register both files (including absent paths), then
run `05` with stage `all`. All Python remains in `04`; graphics caches stay in data.

## Configuration

```{r setup}
options(readr.num_threads = 1L, readr.read_lazy = FALSE)
pipeline_path <- file.path(getwd(), "COMPASS_generate_figures_pipeline.R")
if (!file.exists(pipeline_path)) pipeline_path <- file.path(
  getwd(), "COMPASS", "survival_analysis", "COMPASS_generate_figures_pipeline.R")
pipeline_path <- normalizePath(pipeline_path, mustWork = TRUE)
source(pipeline_path)
source(file.path(dirname(pipeline_path), "figure_data_cache.R"))

env_flag <- function(name, default = FALSE) {
  value <- tolower(Sys.getenv(name, if (default) "true" else "false"))
  if (!value %in% c("true", "false", "1", "0", "yes", "no")) stop(name, " must be true or false")
  value %in% c("true", "1", "yes")
}
env_values <- function(name, default) trimws(strsplit(Sys.getenv(name, default), ",", fixed = TRUE)[[1]])
env_workers <- function(name, default) {
  value <- suppressWarnings(as.integer(Sys.getenv(name, as.character(default))))
  if (is.na(value) || value < 1L) stop(name, " must be a positive integer")
  value
}
NEPC_PROJ_PATH <- Sys.getenv("COMPASS_DATA_ROOT", "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS")
FIG_ROOT <- Sys.getenv("COMPASS_FIG_ROOT", "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS")
FIGURE_DATA_ROOT <- Sys.getenv("COMPASS_FIGURE_DATA_ROOT", file.path(NEPC_PROJ_PATH, "figure_data"))
FIGURE_STAGE <- match.arg(Sys.getenv("COMPASS_FIGURE_STAGE", "all"), c("all", "prepare", "render"))
FIGURE_SCOPE <- match.arg(Sys.getenv("COMPASS_FIGURE_SCOPE", "all"), c("all", "federated"))
# Platinum only by default; set COMPASS_FIGURE_ENDPOINTS=nepc,platinum to restore
# the NEPC landmark figures alongside it (or =nepc for that endpoint alone).
ENDPOINTS <- env_values("COMPASS_FIGURE_ENDPOINTS", "platinum")
ADT_COHORT_OPTIONS <- c("adt", "adt_noprecastrate", "adt_metastatic_adt",
  "adt_metastatic_adt_noprecastrate", "adt_metastatic_llm", "adt_metastatic_llm_noprecastrate")
COHORTS <- env_values("COMPASS_FIGURE_COHORTS", "adt")
stopifnot(length(COHORTS) > 0L, length(ENDPOINTS) > 0L,
          all(COHORTS %in% SUPPORTED_COHORTS), all(ENDPOINTS %in% SUPPORTED_ENDPOINTS),
          !anyDuplicated(COHORTS), !anyDuplicated(ENDPOINTS))
COHORT_FOREST_CONFIG <- list(script = file.path(dirname(pipeline_path), "cohort_forest_figures.R"),
                             cohorts = ADT_COHORT_OPTIONS, landmark = 180L)
PLOT_NON_ANDROGEN_LAB_FIGURES <- env_flag("COMPASS_NON_ANDROGEN_LABS")
PLOT_GAM_TRAJECTORIES <- env_flag("COMPASS_GAM_TRAJECTORIES")
PLOT_ADT_INTENT_SUPPLEMENT <- env_flag("COMPASS_ADT_INTENT_SUPPLEMENT")
PLOT_METASTATIC_LABEL_SUPPLEMENT <- env_flag("COMPASS_METASTATIC_LABEL_SUPPLEMENT", TRUE)
PLOT_FEDERATED_NO_MSK_SUPPLEMENT <- env_flag("COMPASS_FEDERATED_NO_MSK_SUPPLEMENT", TRUE)
METASTATIC_SUPPLEMENT <- list(
  prepare_script = file.path(dirname(pipeline_path), "prepare_metastatic_figure_labels.R"),
  r_script = file.path(dirname(pipeline_path), "metastatic_figure_supplements.R"),
  intent = file.path(NEPC_PROJ_PATH, "mrn_lists", "adt_intent_labels_model_cohort.csv"),
  stage = Sys.getenv("COMPASS_REGEX_STAGE_PATH", file.path(
    Sys.getenv("PROFILE_DATA_PATH", "/data/gusev/USERS/jpconnor/data/PROFILE_DATA"),
    "CANCER_ANNOTATIONS", "CANCER_STAGE_NOTE_LEVEL.parquet")),
  llm = Sys.getenv("LLM_MET_LABELS_PATH", "/data/gusev/USERS/jpconnor/data/LLM_annotations/LLM_met_diagnosis/met_dx_labels.parquet"),
  icd = file.path(NEPC_PROJ_PATH, "prostate_icd_data.csv"))
FEDERATED_NO_MSK_SUPPLEMENT <- list(
  script = file.path(dirname(pipeline_path), "federated_no_msk_figures.R"),
  results = Sys.getenv("COMPASS_FEDERATED_NO_MSK_RESULTS", file.path(
    NEPC_PROJ_PATH, "federated_results_no_MSK", "cox_federated_univariate_adt.csv")))
RENDER_PROFILE <- match.arg(tolower(Sys.getenv("COMPASS_RENDER_PROFILE", "fast")), c("fast", "publication"))
RENDER_DPI <- if (RENDER_PROFILE == "publication") 600 else 200
RENDER_PDF <- env_flag("COMPASS_RENDER_PDF")
RENDER_OVERWRITE <- env_flag("COMPASS_RENDER_OVERWRITE")
PREPARE_OVERWRITE <- env_flag("COMPASS_PREPARE_OVERWRITE")
RENDER_WORKERS <- env_workers("COMPASS_RENDER_WORKERS", 2L)
PREPARE_WORKERS <- env_workers("COMPASS_PREPARE_WORKERS", 1L)
source(file.path(dirname(pipeline_path), "metastatic_figure_supplements.R"))

figure_config <- list(
  data_root = NEPC_PROJ_PATH, cache_root = FIGURE_DATA_ROOT, fig_root = FIG_ROOT,
  cohorts = COHORTS, endpoints = ENDPOINTS, scope = FIGURE_SCOPE,
  labs = if (PLOT_NON_ANDROGEN_LAB_FIGURES) names(CATEGORY_MAP) else ANDROGEN,
  classifier_path = DEFAULT_LLM_ANNOTATIONS_PATH,
  forest_cohorts = COHORT_FOREST_CONFIG$cohorts, forest_landmark = COHORT_FOREST_CONFIG$landmark,
  gam = PLOT_GAM_TRAJECTORIES, adt_intent = PLOT_ADT_INTENT_SUPPLEMENT,
  metastatic = PLOT_METASTATIC_LABEL_SUPPLEMENT, metastatic_extra = metastatic_extra_panels_enabled(),
  metastatic_sources = METASTATIC_SUPPLEMENT[c("intent", "stage", "llm", "icd")],
  federated = PLOT_FEDERATED_NO_MSK_SUPPLEMENT,
  federated_path = FEDERATED_NO_MSK_SUPPLEMENT$results)
```

## Prepare and render

```{r figures}
figure_run <- run_cached_figure_workflow(
  figure_config, pipeline_path, FIGURE_STAGE,
  prepare_workers = PREPARE_WORKERS, render_workers = RENDER_WORKERS,
  dpi = RENDER_DPI, pdf = RENDER_PDF,
  prepare_overwrite = PREPARE_OVERWRITE, render_overwrite = RENDER_OVERWRITE,
  metastatic_config = METASTATIC_SUPPLEMENT,
  forest_config = COHORT_FOREST_CONFIG,
  federated_config = FEDERATED_NO_MSK_SUPPLEMENT)
```
