---
title: "COMPASS endpoint comparison (platinum / NEPC / AVPC)"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

Cross-endpoint comparison for the ADT anchor. Where `05_figures.Rmd` renders
one endpoint's figures at a time, this document puts all three side by side:
event counts, event-positive overlap, pre-anchor events, time to event,
univariate significance overlap, the PSA/Testosterone log-HR panel, and
held-out prediction accuracy per landmark.

**Read-only.** It refits nothing and reads the artifacts written by
`01_preprocessing.ipynb`, `02_univariate.ipynb`, and `03_multivariate.ipynb`
run with `ENDPOINTS = ("platinum", "nepc", "avpc")`. Missing
artifacts are reported in the closing section rather than aborting the knit,
so a partially complete run still renders every section it can.

### Three things to keep in view while reading

**These are not the same patients.** Each endpoint's `prediction_inputs_*`
tree is gated on its own duration being positive, which drops *prevalent*
cases (event at or before the anchor) to make the endpoint incident. Section 2
quantifies that per endpoint and section 3 counts exactly who was dropped.
Any metric difference across endpoints therefore confounds *event* with
*cohort*.

**Sections 2-4 read the Stage 1 cohort, not the per-endpoint trees.** Stage 1
carries all three modeled endpoints' event/duration columns for one common patient set,
before any endpoint-specific gating. That is the only place where the overlap
and pre-anchor questions are answerable: a Venn built from the gated trees
cannot see a pre-anchor event, because the gate already removed it.

**NEPC and AVPC come from the same criteria timeline but are modeled
separately.** The former captures any NEPC feature; the latter captures the
Aparicio-criteria endpoint. Joint AVPC/NEPC fields are audit metadata only.

## Setup

```{r setup, message = FALSE, warning = FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(scales)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5.5, dpi = 150)

NEPC_PROJ_PATH <- "/data/gusev/USERS/jpconnor/data/CAIA/COMPASS"
FIG_ROOT       <- "/data/gusev/USERS/jpconnor/figures/CAIA/COMPASS"

ARM       <- "adt"
LANDMARKS <- c(0, 90, 180)
ID_COL    <- "DFCI_MRN"
ALPHA     <- 0.05

SURVIVAL_ROOT <- file.path(NEPC_PROJ_PATH, "survival_analysis")

# Output for this document only. Kept out of the numbered figure*/ groups that
# generate_figures() manages (and prunes), so nothing here collides with it.
OUT_DIR <- file.path(FIG_ROOT, toupper(ARM), "endpoint_comparison")
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)

# Mirrors COMPASS.survival_analysis.cox_aggregated.ENDPOINTS. The suffix is ""
# for platinum (the historical un-suffixed tree) and "_<endpoint>" otherwise,
# matching compass_pipeline.make_endpoint_runs().
# duration_col is the LANDMARK-relative column produced by make_outcome_df in
# the per-endpoint prediction-input build. stage1_duration_col is the
# ANCHOR-relative column that compile_COMPASS_cohort_data.py writes into the
# Stage 1 cohort. They are different columns at different stages: the lowercase
# t_* names do NOT exist in the Stage 1 CSV, so sections 2-5 (which read Stage 1)
# must use TT_*. Using t_* there silently yields all-NA durations and collapses
# every prevalent/incident count to zero.
ENDPOINT_SPECS <- list(
  platinum   = list(event_col = "PLATINUM",   duration_col = "t_platinum",
                    stage1_duration_col = "TT_PLATINUM",  label = "Platinum"),
  nepc       = list(event_col = "NEPC",       duration_col = "t_nepc",
                    stage1_duration_col = "TT_NEPC",      label = "NEPC"),
  avpc       = list(event_col = "AVPC",       duration_col = "t_avpc",
                    stage1_duration_col = "TT_AVPC",      label = "AVPC")
)
ENDPOINTS <- names(ENDPOINT_SPECS)

for (ep in ENDPOINTS) {
  suffix <- if (identical(ep, "platinum")) "" else paste0("_", ep)
  ENDPOINT_SPECS[[ep]]$inputs  <- file.path(SURVIVAL_ROOT, paste0("prediction_inputs_", ARM, suffix))
  ENDPOINT_SPECS[[ep]]$results <- file.path(SURVIVAL_ROOT, paste0("local_runs_", ARM, suffix))
}

ENDPOINT_LABELS <- vapply(ENDPOINT_SPECS, `[[`, character(1), "label")

# Fixed colorblind-safe assignment, held constant across every panel below.
ENDPOINT_COLORS <- c(
  platinum  = "#2a78d6",
  nepc      = "#eb6834",
  avpc      = "#22a884"
)

# Stage 1 cohort: all three modeled endpoints' columns on one common patient set,
# before any per-endpoint landmark/validity gating.
COHORT_PATH <- file.path(NEPC_PROJ_PATH,
                         sprintf("prostate_%s_survival_cohort_%s.csv", ARM, ARM))

# Collect rather than crash, so a partial run still renders.
MISSING <- character(0)
require_path <- function(path, what) {
  if (file.exists(path)) return(TRUE)
  MISSING <<- c(MISSING, sprintf("%s: %s", what, path))
  FALSE
}

# MRNs arrive as int/float/character across CSVs; normalize before any set op
# or join so a 12345 / 12345.0 round-trip cannot silently break a merge.
normalize_id <- function(x) {
  num <- suppressWarnings(as.numeric(as.character(x)))
  out <- ifelse(is.finite(num), format(num, scientific = FALSE, trim = TRUE), NA_character_)
  trimws(out)
}

save_fig <- function(plot, stem, width = 9, height = 5.5) {
  path <- file.path(OUT_DIR, sprintf("%s_%s.png", ARM, stem))
  ggsave(path, plot = plot, width = width, height = height, units = "in",
         dpi = 600, bg = "white")
  message("wrote ", path)
  invisible(plot)
}

write_out <- function(df, stem) {
  path <- file.path(OUT_DIR, sprintf("%s_%s.csv", ARM, stem))
  readr::write_csv(df, path)
  message("wrote ", path)
  invisible(df)
}

theme_ep <- function(base_size = 11) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title       = element_text(face = "bold", size = base_size + 1),
      plot.subtitle    = element_text(size = base_size - 1, color = "#52514e"),
      axis.text        = element_text(size = base_size - 1.5, color = "#1a1a1a"),
      legend.key       = element_blank(),
      legend.position  = "top",
      strip.background = element_blank(),
      strip.text       = element_text(face = "bold", size = base_size - 1),
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
}
theme_set(theme_ep())

message("data root:   ", NEPC_PROJ_PATH)
message("figure root: ", OUT_DIR)
for (ep in ENDPOINTS) {
  message(sprintf("  %-9s inputs=%-5s results=%-5s", ep,
                  dir.exists(ENDPOINT_SPECS[[ep]]$inputs),
                  dir.exists(ENDPOINT_SPECS[[ep]]$results)))
}
```

## 1. Load the Stage 1 cohort

One row per patient, carrying every endpoint's event flag and duration. All
durations are on the anchor-relative clock (ADT start = day 0), so a value
`<= 0` is an event at or before the anchor.

```{r load-cohort}
cohort <- NULL
if (require_path(COHORT_PATH, "Stage 1 survival cohort")) {
  cohort <- read_csv(COHORT_PATH, show_col_types = FALSE, guess_max = 100000)
  cohort[[ID_COL]] <- normalize_id(cohort[[ID_COL]])
  cohort <- cohort %>% filter(!is.na(.data[[ID_COL]])) %>%
    distinct(.data[[ID_COL]], .keep_all = TRUE)

  present <- vapply(ENDPOINTS, function(ep) {
    ENDPOINT_SPECS[[ep]]$event_col %in% names(cohort)
  }, logical(1))
  cat(sprintf("Stage 1 cohort: %s patients\n", comma(nrow(cohort))))
  cat("endpoint columns present: ",
      paste(sprintf("%s=%s", ENDPOINTS, ifelse(present, "yes", "NO")), collapse = "  "),
      "\n", sep = "")

  # Endpoints whose event column never made it into this cohort build cannot be
  # compared; drop them once here rather than guarding every section.
  ENDPOINTS <- ENDPOINTS[present]
  if (length(ENDPOINTS) < length(ENDPOINT_SPECS))
    warning("Dropping endpoints absent from the Stage 1 cohort: ",
            paste(setdiff(names(ENDPOINT_SPECS), ENDPOINTS), collapse = ", "))
}
```

```{r endpoint-long}
# Long form: one row per (patient, endpoint) with a numeric event flag and the
# anchor-relative duration. Every section below is a summarise() over this.
endpoint_long <- NULL
if (!is.null(cohort) && length(ENDPOINTS) > 0) {
  endpoint_long <- map_dfr(ENDPOINTS, function(ep) {
    spec <- ENDPOINT_SPECS[[ep]]
    # Prefer the Stage 1 (anchor-relative) column; fall back to the
    # landmark-relative name only if a build carried it here instead.
    dur_col <- NULL
    for (cand in c(spec$stage1_duration_col, spec$duration_col)) {
      if (!is.null(cand) && cand %in% names(cohort)) { dur_col <- cand; break }
    }
    if (is.null(dur_col)) {
      # Never fall back to a scalar NA: that silently zeroes every
      # prevalent/incident count instead of reporting the real problem.
      MISSING <<- c(MISSING, sprintf(
        "%s duration column: neither %s nor %s is in the Stage 1 cohort",
        ep, spec$stage1_duration_col, spec$duration_col))
      warning(sprintf("No duration column for %s; its timing rows are dropped.", ep))
      return(NULL)
    }
    tibble(
      !!ID_COL := cohort[[ID_COL]],
      endpoint = ep,
      event    = suppressWarnings(as.numeric(cohort[[spec$event_col]])) %>% replace_na(0),
      duration = suppressWarnings(as.numeric(cohort[[dur_col]])),
      duration_col_used = dur_col
    )
  })

  if (is.null(endpoint_long) || nrow(endpoint_long) == 0) {
    endpoint_long <- NULL
  } else {
    # Report the resolved column per endpoint. If this ever prints a t_* name
    # for a Stage 1 read, the timing columns are not what this document assumes.
    endpoint_long %>%
      distinct(endpoint, duration_col_used) %>%
      { cat("duration column used per endpoint:\n"); print(as.data.frame(.)) }

    # Restrict to endpoints that actually produced rows, so a dropped endpoint
    # does not linger as an empty factor level in every table and plot.
    ENDPOINTS <- ENDPOINTS[ENDPOINTS %in% unique(endpoint_long$endpoint)]
    endpoint_long <- endpoint_long %>%
      mutate(endpoint = factor(endpoint, levels = ENDPOINTS,
                               labels = ENDPOINT_LABELS[ENDPOINTS]))
  }
}
```

## 2. Number of events under each endpoint

Stage 1 counts, before landmark gating. `n_prevalent` are events at or before
the ADT anchor (`duration <= 0`): real events that the incident-endpoint gate
removes, so `n_incident` is what the models can actually see at landmark 0.

```{r event-counts}
event_counts <- NULL
if (!is.null(endpoint_long)) {
  event_counts <- endpoint_long %>%
    group_by(endpoint) %>%
    summarise(
      n_patients      = n(),
      n_events        = sum(event == 1, na.rm = TRUE),
      event_rate_pct  = 100 * mean(event == 1, na.rm = TRUE),
      n_prevalent     = sum(event == 1 & !is.na(duration) & duration <= 0),
      n_incident      = sum(event == 1 & !is.na(duration) & duration > 0),
      n_event_no_time = sum(event == 1 & is.na(duration)),
      .groups = "drop"
    ) %>%
    mutate(pct_of_events_prevalent = 100 * n_prevalent / pmax(n_events, 1))

  write_out(event_counts, "endpoint_event_counts")
  knitr::kable(event_counts, digits = 1)
}
```

```{r event-counts-plot, fig.height = 4.5}
if (!is.null(event_counts)) {
  plot_df <- event_counts %>%
    select(endpoint, n_incident, n_prevalent) %>%
    pivot_longer(-endpoint, names_to = "kind", values_to = "n") %>%
    mutate(kind = factor(kind, levels = c("n_incident", "n_prevalent"),
                         labels = c("Incident (after anchor)", "Prevalent (at/before anchor)")))

  p <- ggplot(plot_df, aes(endpoint, n, fill = kind)) +
    geom_col(width = 0.65) +
    geom_text(aes(label = comma(n)), position = position_stack(vjust = 0.5),
              size = 3.1, color = "white", fontface = "bold") +
    scale_fill_manual(values = c("Incident (after anchor)" = "#2a78d6",
                                 "Prevalent (at/before anchor)" = "#b5b3ae")) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.08))) +
    labs(title = "Event counts by endpoint (Stage 1, ADT anchor)",
         subtitle = "Prevalent events are dropped by each endpoint's incident gate",
         x = NULL, y = "Patients with the event", fill = NULL)
  save_fig(p, "endpoint_event_counts", height = 4.5)
  p
}
```

## 3. Overlap in event-positive patients

Venn-style set statistics over the **event-positive** patients in Stage 1, on
the common patient set (so every endpoint is counted over the same
denominator). Pairwise Jaccard is `|A & B| / |A | B|`.

```{r overlap-pairwise}
positive_sets <- NULL
pairwise <- NULL
if (!is.null(endpoint_long)) {
  positive_sets <- endpoint_long %>%
    filter(event == 1) %>%
    { split(.[[ID_COL]], .$endpoint) } %>%
    lapply(unique)

  pairs <- combn(names(positive_sets), 2, simplify = FALSE)
  pairwise <- map_dfr(pairs, function(pr) {
    a <- positive_sets[[pr[1]]]; b <- positive_sets[[pr[2]]]
    inter <- length(intersect(a, b)); uni <- length(union(a, b))
    tibble(
      endpoint_a = pr[1], endpoint_b = pr[2],
      n_a = length(a), n_b = length(b),
      n_both = inter,
      n_a_only = length(setdiff(a, b)),
      n_b_only = length(setdiff(b, a)),
      jaccard = if (uni > 0) inter / uni else NA_real_,
      pct_of_a_shared = 100 * inter / max(length(a), 1),
      pct_of_b_shared = 100 * inter / max(length(b), 1)
    )
  })
  write_out(pairwise, "endpoint_event_overlap_pairwise")
  knitr::kable(pairwise, digits = c(rep(0, 7), 3, 1, 1))
}
```

```{r overlap-exclusive}
# Full exclusive breakdown: every patient positive for at least one endpoint,
# labeled by the exact combination they fall into. This is the Venn's regions.
combo_counts <- NULL
if (!is.null(endpoint_long)) {
  combo_counts <- endpoint_long %>%
    filter(event == 1) %>%
    group_by(.data[[ID_COL]]) %>%
    summarise(combination = paste(sort(as.character(endpoint)), collapse = " + "),
              n_endpoints = n(), .groups = "drop") %>%
    count(combination, n_endpoints, name = "n_patients") %>%
    arrange(desc(n_patients))

  write_out(combo_counts, "endpoint_event_overlap_combinations")
  knitr::kable(combo_counts)
}
```

```{r overlap-plot, fig.height = 5}
if (!is.null(combo_counts) && nrow(combo_counts) > 0) {
  p <- combo_counts %>%
    mutate(combination = fct_reorder(combination, n_patients)) %>%
    ggplot(aes(n_patients, combination, fill = factor(n_endpoints))) +
    geom_col(width = 0.7) +
    geom_text(aes(label = comma(n_patients)), hjust = -0.15, size = 3.1) +
    scale_fill_brewer(palette = "Blues", name = "Endpoints\nco-occurring") +
    scale_x_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
    labs(title = "Event-positive patients by exact endpoint combination",
         subtitle = "Exclusive regions: each patient is counted once",
         x = "Patients", y = NULL)
  save_fig(p, "endpoint_event_overlap", height = 5)
  p
}
```

## 4. Events occurring before the ADT anchor

Counted as `event == 1 & duration <= 0` on the anchor-relative clock. These
are the prevalent cases each endpoint's gate removes. A large share here means
the incident cohort is a materially different population from Stage 1, and it
is also the mechanism behind the cohort-size differences in section 8.

```{r pre-adt}
pre_adt <- NULL
if (!is.null(endpoint_long)) {
  pre_adt <- endpoint_long %>%
    filter(event == 1, !is.na(duration)) %>%
    group_by(endpoint) %>%
    summarise(
      n_events            = n(),
      n_before_anchor     = sum(duration <= 0),
      pct_before_anchor   = 100 * mean(duration <= 0),
      median_days_before  = median(duration[duration <= 0], na.rm = TRUE),
      earliest_days_before = suppressWarnings(min(duration[duration <= 0], na.rm = TRUE)),
      n_after_anchor      = sum(duration > 0),
      .groups = "drop"
    ) %>%
    mutate(across(where(is.numeric), ~ ifelse(is.finite(.x), .x, NA_real_)))

  write_out(pre_adt, "endpoint_events_before_adt")
  knitr::kable(pre_adt, digits = 1)
}
```

```{r pre-adt-plot, fig.height = 4.5}
if (!is.null(endpoint_long)) {
  dens_df <- endpoint_long %>% filter(event == 1, !is.na(duration))
  if (nrow(dens_df) > 0) {
    p <- ggplot(dens_df, aes(duration, fill = endpoint, color = endpoint)) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#52514e") +
      geom_density(alpha = 0.25, adjust = 0.9) +
      scale_fill_manual(values = unname(ENDPOINT_COLORS[ENDPOINTS])) +
      scale_color_manual(values = unname(ENDPOINT_COLORS[ENDPOINTS])) +
      scale_x_continuous(labels = comma) +
      labs(title = "Event timing relative to the ADT anchor",
           subtitle = "Mass left of the dashed line is prevalent disease, dropped by the incident gate",
           x = "Days from ADT start (negative = before ADT)", y = "Density",
           fill = NULL, color = NULL)
    save_fig(p, "endpoint_event_timing_density", height = 4.5)
    p
  }
}
```

## 5. Time to event

Restricted to incident events (`duration > 0`), since a prevalent event's
negative duration is not a "time to event". Mean is reported because it was
asked for; median is the one to quote in text, as these distributions are
right-skewed and the mean is sensitive to the administrative tail.

```{r time-to-event}
tte <- NULL
if (!is.null(endpoint_long)) {
  tte <- endpoint_long %>%
    filter(event == 1, !is.na(duration), duration > 0) %>%
    group_by(endpoint) %>%
    summarise(
      n_incident_events = n(),
      mean_days   = mean(duration),
      sd_days     = sd(duration),
      median_days = median(duration),
      q1_days     = quantile(duration, 0.25),
      q3_days     = quantile(duration, 0.75),
      min_days    = min(duration),
      max_days    = max(duration),
      mean_years  = mean(duration) / 365.25,
      .groups = "drop"
    )

  write_out(tte, "endpoint_time_to_event")
  knitr::kable(tte, digits = 1)
}
```

```{r time-to-event-plot, fig.height = 4.5}
if (!is.null(endpoint_long)) {
  box_df <- endpoint_long %>% filter(event == 1, !is.na(duration), duration > 0)
  if (nrow(box_df) > 0) {
    p <- ggplot(box_df, aes(endpoint, duration, fill = endpoint)) +
      geom_boxplot(width = 0.55, outlier.alpha = 0.25, outlier.size = 0.8) +
      stat_summary(fun = mean, geom = "point", shape = 23, size = 2.4,
                   fill = "white", color = "#1a1a1a") +
      scale_fill_manual(values = unname(ENDPOINT_COLORS[ENDPOINTS]), guide = "none") +
      scale_y_continuous(labels = comma) +
      labs(title = "Time from ADT start to event, incident events only",
           subtitle = "Diamond = mean, line = median",
           x = NULL, y = "Days to event")
    save_fig(p, "endpoint_time_to_event", height = 4.5)
    p
  }
}
```

## 6. Univariate significance: shared vs distinct vs total

Reads each endpoint's `cox_agg_univariate_nobs_adjusted.csv` per landmark and
counts nominally significant features (`p < 0.05`), then the same on the
BH-adjusted `q_value` when present.

A feature significant for one endpoint and not another is **not** evidence of
a real difference when the two arms have very different event counts —
section 2 sets that expectation, and `n_events_used` is carried through here
so the comparison can be read against it.

```{r load-univariate}
univariate <- NULL
uni_frames <- list()
for (ep in ENDPOINTS) {
  for (lm in LANDMARKS) {
    path <- file.path(ENDPOINT_SPECS[[ep]]$results, "cox",
                      sprintf("landmark_%s", lm), "both",
                      "cox_agg_univariate_nobs_adjusted.csv")
    if (!require_path(path, sprintf("%s univariate landmark %s", ep, lm))) next
    frame <- read_csv(path, show_col_types = FALSE)
    # Each tree models one endpoint, but the CSV carries an endpoint column;
    # filter to this run's endpoint so a stale extra row cannot leak in.
    if ("endpoint" %in% names(frame))
      frame <- frame %>% filter(tolower(as.character(endpoint)) == ep)
    if (nrow(frame) == 0) next
    frame$endpoint <- ep
    frame$landmark_days <- lm
    uni_frames[[length(uni_frames) + 1]] <- frame
  }
}
if (length(uni_frames) > 0) {
  univariate <- bind_rows(uni_frames) %>%
    mutate(
      p_value = suppressWarnings(as.numeric(p_value)),
      q_value = if ("q_value" %in% names(.)) suppressWarnings(as.numeric(q_value)) else NA_real_,
      coef_feature = if ("coef_feature" %in% names(.)) suppressWarnings(as.numeric(coef_feature)) else NA_real_,
      sig_nominal = !is.na(p_value) & p_value < ALPHA,
      sig_fdr     = !is.na(q_value) & q_value < ALPHA
    )
  cat(sprintf("univariate rows: %s across %d endpoint(s) x %d landmark(s)\n",
              comma(nrow(univariate)),
              n_distinct(univariate$endpoint), n_distinct(univariate$landmark_days)))
}
```

```{r univariate-totals}
uni_totals <- NULL
if (!is.null(univariate)) {
  uni_totals <- univariate %>%
    group_by(endpoint, landmark_days) %>%
    summarise(
      n_features_tested = n_distinct(feature),
      n_fitted          = sum(!is.na(p_value)),
      n_sig_nominal     = n_distinct(feature[sig_nominal]),
      n_sig_fdr         = n_distinct(feature[sig_fdr]),
      median_events_used = if ("n_events_used" %in% names(univariate))
        median(suppressWarnings(as.numeric(n_events_used)), na.rm = TRUE) else NA_real_,
      .groups = "drop"
    ) %>%
    arrange(landmark_days, endpoint)

  write_out(uni_totals, "univariate_significance_totals")
  knitr::kable(uni_totals, digits = 1)
}
```

```{r univariate-shared}
# Shared vs distinct, per landmark. "shared_all" = significant under every
# endpoint that has a fitted result for that feature at that landmark;
# "unique_to" = significant under exactly one.
uni_overlap <- NULL
if (!is.null(univariate)) {
  sig_sets <- univariate %>%
    filter(sig_nominal) %>%
    distinct(landmark_days, endpoint, feature)

  uni_overlap <- sig_sets %>%
    group_by(landmark_days, feature) %>%
    summarise(
      n_endpoints = n_distinct(endpoint),
      endpoints   = paste(sort(unique(endpoint)), collapse = " + "),
      .groups = "drop"
    )

  overlap_summary <- uni_overlap %>%
    group_by(landmark_days) %>%
    summarise(
      n_sig_any        = n_distinct(feature),
      n_shared_all     = sum(n_endpoints == length(ENDPOINTS)),
      n_shared_2plus   = sum(n_endpoints >= 2),
      n_unique_to_one  = sum(n_endpoints == 1),
      .groups = "drop"
    )
  write_out(overlap_summary, "univariate_shared_distinct_summary")
  knitr::kable(overlap_summary)

  combo_summary <- uni_overlap %>%
    count(landmark_days, endpoints, n_endpoints, name = "n_features") %>%
    arrange(landmark_days, desc(n_features))
  write_out(combo_summary, "univariate_significance_combinations")
  knitr::kable(combo_summary)
}
```

```{r univariate-pairwise}
uni_pairwise <- NULL
if (!is.null(univariate)) {
  uni_pairwise <- map_dfr(LANDMARKS, function(lm) {
    sets <- univariate %>%
      filter(landmark_days == lm, sig_nominal) %>%
      { split(.$feature, .$endpoint) } %>%
      lapply(unique)
    if (length(sets) < 2) return(tibble())
    map_dfr(combn(names(sets), 2, simplify = FALSE), function(pr) {
      a <- sets[[pr[1]]]; b <- sets[[pr[2]]]
      uni <- length(union(a, b))
      tibble(landmark_days = lm, endpoint_a = pr[1], endpoint_b = pr[2],
             n_a = length(a), n_b = length(b), n_shared = length(intersect(a, b)),
             n_a_only = length(setdiff(a, b)), n_b_only = length(setdiff(b, a)),
             jaccard = if (uni > 0) length(intersect(a, b)) / uni else NA_real_)
    })
  })
  if (nrow(uni_pairwise) > 0) {
    write_out(uni_pairwise, "univariate_significance_pairwise")
    knitr::kable(uni_pairwise, digits = 3)
  }
}
```

```{r univariate-plot, fig.height = 4.5}
if (!is.null(uni_totals)) {
  p <- uni_totals %>%
    select(endpoint, landmark_days, n_sig_nominal, n_sig_fdr) %>%
    pivot_longer(starts_with("n_sig"), names_to = "threshold", values_to = "n") %>%
    mutate(threshold = factor(threshold, levels = c("n_sig_nominal", "n_sig_fdr"),
                              labels = c("nominal p < 0.05", "FDR q < 0.05")),
           endpoint = factor(endpoint, levels = ENDPOINTS,
                             labels = ENDPOINT_LABELS[ENDPOINTS])) %>%
    ggplot(aes(factor(landmark_days), n, fill = endpoint)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.72) +
    facet_wrap(~ threshold) +
    scale_fill_manual(values = unname(ENDPOINT_COLORS[ENDPOINTS]), name = NULL) +
    labs(title = "Significant univariate lab features by endpoint and landmark",
         subtitle = "Counts are sensitive to each endpoint's event count; read with section 2",
         x = "Landmark (days from ADT start)", y = "Significant features")
  save_fig(p, "univariate_significance_counts", height = 4.5)
  p
}
```

## 7. PSA and Testosterone: direction, magnitude, significance

`coef_feature` is the log hazard ratio per SD of the feature, so its **sign is
the direction**: positive = higher values associate with shorter time to
event. The 95% CI is reported on the HR scale by the pipeline and is
log-transformed here so the interval matches the plotted point estimate.

Features are matched on the canonical lab name, following the R pipeline's
convention that the raw OMOP `Prostate specific Ag` maps to PSA. All summary
statistics of each lab (mean, slope, last value, ...) are kept and shown as
separate rows, since direction can legitimately differ between a level and a
trend feature.

```{r androgen-extract}
ANDROGEN <- c("PSA", "Testosterone")

# Mirrors match_canonical_lab_name() in COMPASS_generate_figures_pipeline.R.
match_androgen <- function(x) {
  n <- tolower(as.character(x))
  out <- rep(NA_character_, length(n))
  out[grepl("prostate specific ag", n, fixed = FALSE)] <- "PSA"
  out[is.na(out) & grepl("^psa$|(^|[^a-z])psa([^a-z]|$)", n)] <- "PSA"
  out[is.na(out) & grepl("testosterone", n)] <- "Testosterone"
  out
}

androgen <- NULL
if (!is.null(univariate)) {
  # feature is "<lab>__<stat>"; prefer the pipeline's own lab_name/feature_stat
  # columns when present and fall back to splitting the feature name.
  androgen <- univariate %>%
    mutate(
      .lab  = if ("lab_name" %in% names(.)) lab_name else str_replace(feature, "__[^_]*$", ""),
      .stat = if ("feature_stat" %in% names(.)) feature_stat
              else ifelse(str_detect(feature, "__"), str_extract(feature, "[^_]+$"), "value"),
      androgen_lab = match_androgen(.lab)
    ) %>%
    filter(!is.na(androgen_lab)) %>%
    mutate(
      log_hr    = coef_feature,
      hr        = suppressWarnings(as.numeric(hazard_ratio_per_sd)),
      log_ci_lo = suppressWarnings(log(as.numeric(ci_lower))),
      log_ci_hi = suppressWarnings(log(as.numeric(ci_upper))),
      direction = case_when(is.na(log_hr) ~ NA_character_,
                            log_hr > 0 ~ "higher -> shorter time to event",
                            log_hr < 0 ~ "higher -> longer time to event",
                            TRUE ~ "null"),
      significance = case_when(sig_fdr ~ "FDR q < 0.05",
                               sig_nominal ~ "nominal p < 0.05",
                               TRUE ~ "ns")
    )

  if (nrow(androgen) == 0) {
    MISSING <- c(MISSING, "PSA/Testosterone features: none matched in the univariate results")
  } else {
    androgen_table <- androgen %>%
      select(androgen_lab, source_lab = .lab, feature_stat = .stat,
             endpoint, landmark_days,
             log_hr, hr, log_ci_lo, log_ci_hi, p_value, q_value,
             any_of(c("n_patients_used", "n_events_used")),
             direction, significance) %>%
      arrange(androgen_lab, source_lab, feature_stat, landmark_days, endpoint)
    write_out(androgen_table, "androgen_loghr_by_endpoint")
    knitr::kable(androgen_table, digits = 3)
  }
}
```

```{r androgen-plot, fig.height = 7}
if (!is.null(androgen) && nrow(androgen) > 0) {
  plot_df <- androgen %>%
    filter(!is.na(log_hr)) %>%
    mutate(endpoint = factor(endpoint, levels = ENDPOINTS,
                             labels = ENDPOINT_LABELS[ENDPOINTS]),
           # Facet on the source lab so an aliased variant ("Prostate specific
           # Ag" vs "PSA") gets its own panel instead of overplotting.
           panel = ifelse(.lab == androgen_lab,
                          sprintf("%s :: %s", androgen_lab, .stat),
                          sprintf("%s [%s] :: %s", androgen_lab, .lab, .stat)))

  if (nrow(plot_df) > 0) {
    p <- ggplot(plot_df, aes(log_hr, endpoint, color = endpoint, shape = significance)) +
      geom_vline(xintercept = 0, linetype = "dashed", color = "#52514e") +
      geom_errorbar(aes(xmin = log_ci_lo, xmax = log_ci_hi),
                    orientation = "y", width = 0.18,
                    linewidth = 0.5, na.rm = TRUE) +
      geom_point(size = 2.4) +
      facet_grid(panel ~ landmark_days, scales = "free_y",
                 labeller = labeller(landmark_days = function(x) paste0("landmark +", x, "d"))) +
      scale_color_manual(values = unname(ENDPOINT_COLORS[ENDPOINTS]), guide = "none") +
      scale_shape_manual(values = c("FDR q < 0.05" = 16, "nominal p < 0.05" = 17, "ns" = 1),
                         name = NULL) +
      labs(title = "PSA and Testosterone log hazard ratio per SD, by endpoint",
           subtitle = "Right of the dashed line = higher values associate with shorter time to event; bars are 95% CI",
           x = "log HR per SD", y = NULL)
    save_fig(p, "androgen_loghr_forest", height = max(5, 1.1 * n_distinct(plot_df$panel) + 2))
    p
  }
}
```

```{r androgen-direction}
# Compact direction/significance grid: the "does PSA point the same way for
# every endpoint" question, answered without reading the forest plot.
if (!is.null(androgen) && nrow(androgen) > 0) {
  # Key on the SOURCE lab name, not the canonical one. Several raw lab names
  # (e.g. "Prostate specific Ag" and "PSA") collapse to the same canonical lab
  # but are distinct columns in the feature matrix; keying on the canonical
  # name would make the key non-unique, and pivot_wider would silently emit
  # list-columns that serialize to blank cells.
  direction_grid <- androgen %>%
    filter(!is.na(log_hr)) %>%
    mutate(cell = sprintf("%+.3f%s", log_hr,
                          case_when(sig_fdr ~ "**", sig_nominal ~ "*", TRUE ~ ""))) %>%
    select(androgen_lab, source_lab = .lab, feature_stat = .stat,
           landmark_days, endpoint, cell)

  dup_keys <- direction_grid %>%
    count(androgen_lab, source_lab, feature_stat, landmark_days, endpoint) %>%
    filter(n > 1)
  if (nrow(dup_keys) > 0) {
    warning("Duplicate (lab, stat, landmark, endpoint) keys in the androgen grid; ",
            "showing the first fitted value per key. Inspect ",
            "androgen_loghr_by_endpoint.csv for the full set.")
    direction_grid <- direction_grid %>%
      group_by(androgen_lab, source_lab, feature_stat, landmark_days, endpoint) %>%
      slice(1) %>% ungroup()
  }

  direction_grid <- direction_grid %>%
    pivot_wider(names_from = endpoint, values_from = cell) %>%
    arrange(androgen_lab, source_lab, feature_stat, landmark_days)

  write_out(direction_grid, "androgen_loghr_direction_grid")
  cat("log HR per SD;  * nominal p < 0.05,  ** FDR q < 0.05\n\n")
  knitr::kable(direction_grid)
}
```

## 8. Prediction accuracy at each timepoint

Held-out test metrics only, never cross-validation, read from the same
per-model metrics files `compass_pipeline.summarize_outputs()` reads. Both
`both` (full feature set) and `baseline` configs are included so the
lab-feature contribution is visible per endpoint.

Differences across endpoints confound event definition with cohort and with
event count — `n_test_events` is carried in every table for that reason.

```{r load-performance}
MODEL_SPECS <- tribble(
  ~model,         ~subdir,   ~config,    ~metrics_file,
  "elastic-net",  "cox",     "both",     "cox_agg_multivariable_metrics.csv",
  "elastic-net",  "cox",     "baseline", "cox_agg_baseline_metrics.csv",
  "xgboost",      "xgboost", "both",     "landmark_xgboost_metrics.csv",
  "xgboost",      "xgboost", "baseline", "landmark_xgboost_baseline_metrics.csv"
)

read_metrics_row <- function(path, ep, model) {
  df <- read_csv(path, show_col_types = FALSE)
  if ("endpoint" %in% names(df))
    df <- df %>% filter(tolower(as.character(endpoint)) == ep)
  if (nrow(df) == 0) return(NULL)
  row <- df[1, ]
  # elastic-net and xgboost emit the same metrics under different column names.
  pick <- function(...) {
    for (nm in c(...)) if (nm %in% names(row)) return(suppressWarnings(as.numeric(row[[nm]])))
    NA_real_
  }
  tibble(
    n_test          = pick("n_test"),
    n_test_events   = pick("n_events_test", "n_test_events"),
    c_index         = pick("test_c_index", "c_index"),
    mean_auc_t      = pick("test_mean_auc_t", "mean_auc_t"),
    integrated_brier = pick("test_integrated_brier", "integrated_brier")
  )
}

performance <- NULL
perf_rows <- list()
for (ep in ENDPOINTS) {
  for (i in seq_len(nrow(MODEL_SPECS))) {
    spec <- MODEL_SPECS[i, ]
    for (lm in LANDMARKS) {
      path <- file.path(ENDPOINT_SPECS[[ep]]$results, spec$subdir,
                        sprintf("landmark_%s", lm), spec$config, spec$metrics_file)
      if (!require_path(path, sprintf("%s %s/%s landmark %s", ep, spec$model, spec$config, lm)))
        next
      vals <- read_metrics_row(path, ep, spec$model)
      if (is.null(vals)) {
        MISSING <- c(MISSING, sprintf("%s %s/%s landmark %s: no %s row in %s",
                                      ep, spec$model, spec$config, lm, ep, path))
        next
      }
      perf_rows[[length(perf_rows) + 1]] <- bind_cols(
        tibble(endpoint = ep, model = spec$model, config = spec$config, landmark = lm),
        vals
      )
    }
  }
}
if (length(perf_rows) > 0) {
  performance <- bind_rows(perf_rows) %>% arrange(endpoint, model, config, landmark)
  write_out(performance, "endpoint_performance")
  knitr::kable(performance, digits = 3)
}
```

```{r performance-pivots}
if (!is.null(performance)) {
  for (metric in c("c_index", "mean_auc_t", "integrated_brier")) {
    wide <- performance %>%
      select(model, config, landmark, endpoint, all_of(metric)) %>%
      pivot_wider(names_from = endpoint, values_from = all_of(metric)) %>%
      arrange(model, config, landmark)
    # Every non-platinum endpoint's delta against platinum, the reference arm.
    if ("platinum" %in% names(wide)) {
      for (ep in setdiff(ENDPOINTS, "platinum")) {
        if (ep %in% names(wide))
          wide[[paste0(ep, " - platinum")]] <- wide[[ep]] - wide[["platinum"]]
      }
    }
    write_out(wide, sprintf("endpoint_performance_%s", metric))
    cat(sprintf("\n\n### %s\n\n", metric))
    print(knitr::kable(wide, digits = 3))
  }
}
```

```{r performance-plot, fig.height = 6.5}
if (!is.null(performance)) {
  plot_df <- performance %>%
    select(endpoint, model, config, landmark, c_index, mean_auc_t, integrated_brier) %>%
    pivot_longer(c(c_index, mean_auc_t, integrated_brier),
                 names_to = "metric", values_to = "value") %>%
    filter(!is.na(value)) %>%
    mutate(endpoint = factor(endpoint, levels = ENDPOINTS,
                             labels = ENDPOINT_LABELS[ENDPOINTS]),
           arm = paste(model, config, sep = " / "))

  if (nrow(plot_df) > 0) {
    p <- ggplot(plot_df, aes(landmark, value, color = endpoint, group = endpoint)) +
      geom_hline(data = tibble(metric = "c_index", y = 0.5), aes(yintercept = y),
                 linetype = "dashed", color = "#52514e") +
      geom_hline(data = tibble(metric = "mean_auc_t", y = 0.5), aes(yintercept = y),
                 linetype = "dashed", color = "#52514e") +
      geom_line(linewidth = 0.6) +
      geom_point(size = 2.2) +
      facet_grid(metric ~ arm, scales = "free_y") +
      scale_x_continuous(breaks = LANDMARKS) +
      scale_color_manual(values = unname(ENDPOINT_COLORS[ENDPOINTS]), name = NULL) +
      labs(title = "Held-out prediction accuracy by endpoint and landmark",
           subtitle = "Dashed line = chance for discrimination metrics; lower is better for integrated Brier",
           x = "Landmark (days from ADT start)", y = NULL)
    save_fig(p, "endpoint_performance", height = 6.5)
    p
  }
}
```

```{r performance-events}
# Event counts behind each metric. The interpretive guardrail for section 8:
# a lower C-index on a few dozen test events is a statement about power.
if (!is.null(performance)) {
  event_support <- performance %>%
    distinct(endpoint, model, config, landmark, n_test, n_test_events) %>%
    arrange(endpoint, landmark, model, config)
  write_out(event_support, "endpoint_performance_event_support")

  thin <- event_support %>% filter(!is.na(n_test_events), n_test_events < 25)
  if (nrow(thin) > 0) {
    cat("!! arms with fewer than 25 held-out events -- treat their metrics as underpowered:\n")
    print(knitr::kable(thin))
  }
}
```

## 9. Missing artifacts

```{r missing}
if (length(MISSING) > 0) {
  cat(sprintf("%d artifact(s) were not found:\n\n", length(unique(MISSING))))
  for (item in unique(MISSING)) cat("  ", item, "\n", sep = "")
  cat("\nRun 01/02/03 for the corresponding endpoint with",
      "ENDPOINTS = (\"platinum\", \"nepc\", \"avpc\").\n")
} else {
  cat("All expected artifacts were found.\n")
}
```

```{r session-info}
sessionInfo()
```
